---
title: "Feather River Mini Snorkel Data QC"
author: "Maddee Rubenson"
date: "August 21, 2024"
output: 
  html_document:
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(googleCloudStorageR)
library(ggplot2)
library(scales)

colors_small <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                   "#899DA4", "#C93312", "#DC863B" # royal 1 (- 3)
)
```

# Feather River Mini Snorkel Data - 2001 & 2002

## Description of Monitoring Data

This markdown explores the data that was provided for mini snorkel data
for Feather River in the database: `minisnorkdb3.mdb`. Previously, a
database was provided that only contained 2001 data
(`MiniSnorkelDTB.mdb`). Upon exploration and comparison between this new
database and the old database, it was decided that we will move forward
using this database for publishing on EDI. For details on the
`MiniSnorkelDTB` database visit: `qc/qc_mini_snorkel_data.Rmd` and for
details on the database comparison, visit: `qc/database_qc.Rmd`.

**Timeframe:** March 2001 - August 2002

**Completeness of Record throughout timeframe:** fairly complete

**Sampling Location:** Feather River

**Data Contact:** Ryon Kurth

## Source Database pull

-   Database name: `minisnorkdb3.mdb`

```{r include=FALSE}
source(here::here('edi-mini-snorkel','data-raw', 'query_4mac.R'))
```

Read in data sourced from query script, glimpse raw data and domain
description sheet:

```{r}
# read in data to clean 
microhabitat_2002 |> glimpse()
```

## Data transformations

### first table reviewed is the All Fish Observation Table

All of the substrate and cover lookups are not true look ups. Substrate
and cover column indicate a percentage of cover or substrate of each
type. I updated the column names to reflect this and utilized the lookup
tables to understand which substrate or cover type each column was
referring to.

Columns removed:

-   SpecAge removed - just a combination of species and age

-   i_cov_sum - removed because sum of other columns

-   o_cov_sum - removed because sum of other columns

-   sub_sum - removed because sum of other columns

-   start_time - just a date that seemed wrong

-   end_time - also just a date that seemed wrong

-   crew - specific crew names do not need to be present on public EDI
dataset

```{r}
# Combine the data here
# 2001/2002 data
all_fish_data <- microhabitat_2002 |> 
  left_join(fish_data_2002, by = c("TCode" = "TCode", "PDatID" = "PDatID")) |> 
  left_join(species_code_lookup_2002, by = c("Species" = "SpeciesCodeID")) |> # all codes are in the lookup
  select(-c(SpeciesCode, Species)) |>
  rename(species = Species.y, 
         percent_fine_substrate = Sub1, 
         percent_sand_substrate = Sub2, 
         percent_small_gravel_substrate = Sub3, 
         percent_large_gravel_substrate = Sub4, 
         percent_cobble_substrate = Sub5, 
         percent_boulder_substrate = Sub6,
         percent_no_cover_inchannel = IcovA, 
         percent_small_woody_cover_inchannel = IcovB, 
         percent_large_woody_cover_inchannel = IcovC, 
         percent_submerged_aquatic_veg_inchannel = IcovE, 
         percent_undercut_bank = IcovF,
         percent_no_cover_overhead = Ocov0,
         percent_cover_half_meter_overhead = Ocov1, 
         percent_cover_more_than_half_meter_overhead = Ocov2)|> 
  # Clean in separate file
  left_join(location_table_2002, by = c("PDatID" = "PhysDataTblID")) |> 
  left_join(weather_code_lookup_2002, by = c("Weather" = "WeatherCodeLookUpID")) |> # all codes are in the lookup
  select(-WeatherCode, -Weather) |> 
  rename(weather = Weather.y) |> 
  # note that there are 0 channel types which do not map to the lookup
  left_join(channel_lookup_2002, by = c("ChannelType" = "ChannelTypeCode")) |> 
  select(-ChannelType, -ChannelTypeCodeID) |> 
  rename(channel_type = ChannelType.y) |> 
  janitor::clean_names() |> 
  # fixes issues with the codes so the CGU lookup will work
  mutate(cgu = tolower(cgu),
         cgu = case_when(cgu == "rm`" ~ "rm",
                         cgu == "gm." ~ "gm",
                         cgu == "" ~ NA,
                         T ~ cgu)) |> 
  left_join(cgu_code_lookup_2002 |> mutate(CGUCode = tolower(CGUCode)), by = c("cgu" = "CGUCode")) |> 
  select(-cgu, -CGUCodeID, -sub_sum, -i_cov_sum, -o_cov_sum, -start_time, -end_time, -crew) |> 
  rename(channel_geomorphic_unit = CGU) |> 
  filter(!is.na(date)) |> 
  write_csv('2002_fish_obs_tmp.csv') # annoying workaround to remove labeled date column 

all_fish_data <- read_csv('2002_fish_obs_tmp.csv')  |> glimpse()

```

**NA and Unknown Date Values**

There are `r is.na(all_fish_data$date) |> sum()` NA values

## Explore Numeric Variables: {.tabset}

```{r}
# Filter clean data to show only numeric variables 
all_fish_data %>% select_if(is.numeric) %>% colnames()

```

### Variable: `count`

**Plotting Count over Period of Record**

```{r}
# Make whatever plot is appropriate 
# maybe 2+ plots are appropriate
all_fish_data %>% 
  ggplot(aes(x = rm, y = count, group = 1))+
  # geom_line()+
  geom_point(aes(x=date, y = count))+
  theme_minimal()+
  labs(title = "Count over Time",
       y = "Number of Fish Observations")
```

**Numeric Summary of Count over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$count)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$count) |> sum()` NA values

### Variable: `date`

**Plotting Date over Period of Record**

All observations are from 2001

```{r}
# Make whatever plot is appropriate 
# maybe 2+ plots are appropriate
all_fish_data %>% 
  ggplot(aes(x = date, y = TRUE, group = 1))+
  # geom_line()+
  geom_point(aes(x=date, y = TRUE))+
  theme_minimal()+
  labs(title = "Dates Surveyed",
       y = "")
```

**Numeric Summary of Count over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$date)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$date) |> sum()` NA values

### Variable: \`river_mile\`\`

**Plotting river_mile over Period of Record**

```{r}
# Make whatever plot is appropriate 
# maybe 2+ plots are appropriate
all_fish_data %>% 
  ggplot(aes(x = river_mile, y = date, group = 1))+
  # geom_line()+
  geom_point(aes(x=river_mile, y = date))+
  theme_minimal()+
  labs(title = "Dates that River miles Surveyed",
       y = "")
```

**Numeric Summary of river_mile over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$river_mile)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$river_mile) |> sum()` NA values

### Variable: `fl_mm`

**Plotting fl_mm over Period of Record**

```{r}
# Make whatever plot is appropriate 
# maybe 2+ plots are appropriate
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = fl_mm)) + 
  geom_histogram(breaks=seq(0, 200, by=2)) + 
  scale_x_continuous(breaks=seq(0, 200, by=25)) +
  theme_minimal() +
  labs(title = "Fork length distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

```{r}
all_fish_data %>% 
  mutate(year = as.factor(year(date))) %>%
  ggplot(aes(x = fl_mm, y = species)) + 
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Fork length summarized by species") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

**Numeric Summary of fl_mm over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$fl_mm)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$fl_mm) |> sum()` NA values

### Variable: `dist_to_bottom`

**Plotting dist_to_bottom distribution**

```{r}
# Make whatever plot is appropriate 
# maybe 2+ plots are appropriate
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = dist_to_bottom)) + 
  geom_histogram(breaks=seq(0, 200, by=2)) + 
  scale_x_continuous(breaks=seq(0, 200, by=25)) +
  theme_minimal() +
  labs(title = "Distance to Bottom") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

```{r}

all_fish_data %>% 
  mutate(year = as.factor(year(date))) %>%
  ggplot(aes(x = dist_to_bottom, y = as.factor(river_mile))) + 
  geom_boxplot() + 
  theme_minimal() +
  labs(title = "Distance to bottom summarized by river_mile") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of dist_to_bottom over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$river_mile)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$river_mile) |> sum()` NA values

### Variable: `focal_velocity` & `velocity`

**Plotting velocities over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = focal_velocity)) + 
  geom_histogram(breaks=seq(0, 10, by=1), group = "Focal velocity", fill = "blue", alpha = .2 ) + 
  geom_histogram(aes(x = velocity), breaks=seq(0, 10, by=1), group = "Focal velocity", fill = "red", alpha = .2 ) + 
  scale_x_continuous(breaks=seq(0, 10, by=1)) +
  theme_minimal() +
  labs(title = "Focal Velocity vs ") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

Looks like velocity and focal velocity have similar distributions

**Numeric Summary of focal_velocity over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$focal_velocity)
summary(all_fish_data$velocity)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$focal_velocity) |> sum()` NA values
There are `r is.na(all_fish_data$velocity) |> sum()` NA values

### Variable: `t_code`

**Plotting t_code over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = t_code)) + 
  geom_histogram(breaks=seq(0, 30, by=1)) + 
  scale_x_continuous(breaks=seq(0, 30, by=1)) +
  theme_minimal() +
  labs(title = "T code distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of t_code over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$t_code)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$t_code) |> sum()` NA values

### Variable: `depth`

**Plotting depth over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = depth)) + 
  geom_histogram(breaks=seq(0, 250, by=1)) + 
  scale_x_continuous(breaks=seq(0, 200, by=50)) +
  theme_minimal() +
  labs(title = "T code distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of depth over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$depth)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$t_code) |> sum()` NA values

### Variable: `substrate`

**Plotting substrate over Period of Record**

```{r}
# Make whatever plot is appropriate 
# maybe 2+ plots are appropriate
all_fish_data |> 
  arrange(river_mile) |> 
  mutate(obs_id = 1:nrow(all_fish_data)) |> 
  select(obs_id, 
         percent_fine_substrate,
         percent_sand_substrate,
         percent_small_gravel_substrate,
         percent_large_gravel_substrate,
         percent_cobble_substrate,
         percent_boulder_substrate) |> 
  pivot_longer(cols = percent_fine_substrate:percent_boulder_substrate, 
               names_to = "substrate_type", values_to = "percent") |> 
  ggplot(aes(x = obs_id, y = percent, fill = substrate_type)) +
  geom_col() +
  scale_fill_manual(values = colors_small) +
  theme_minimal() +
  theme(legend.position = "bottom")


```

**Numeric Summary of substrate over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$percent_fine_substrate)
summary(all_fish_data$percent_sand_substrate)
summary(all_fish_data$percent_small_gravel_substrate)
summary(all_fish_data$percent_large_gravel_substrate)
summary(all_fish_data$percent_boulder_substrate)
summary(all_fish_data$percent_cobble_substrate)

```

**NA and Unknown Values**

There are `r is.na(all_fish_data$percent_fine_substrate) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_sand_substrate) |> sum()` NA values There
are `r is.na(all_fish_data$percent_small_gravel_substrate) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_small_gravel_substrate) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_boulder_substrate) |> sum()` NA values
There are `r is.na(all_fish_data$percent_cobble_substrate) |> sum()` NA
values

### Variable: `inchannel cover`

**Plotting inchannel cover over Period of Record**

Notes: - some cover totals less than 100%

```{r}
all_fish_data |> 
  arrange(river_mile) |> 
  mutate(obs_id = 1:nrow(all_fish_data)) |> 
  select(obs_id, 
         percent_no_cover_inchannel,
         percent_small_woody_cover_inchannel,
         percent_large_woody_cover_inchannel,
         percent_submerged_aquatic_veg_inchannel) |> 
  pivot_longer(cols = percent_no_cover_inchannel:percent_submerged_aquatic_veg_inchannel, 
               names_to = "inchannel_cover_type", values_to = "percent") |> 
  ggplot(aes(x = obs_id, y = percent, fill = inchannel_cover_type)) +
  geom_col() +
  scale_fill_manual(values = colors_small[4:7]) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Numeric Summary of inchannel cover over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$percent_no_cover_inchannel)
summary(all_fish_data$percent_small_woody_cover_inchannel)
summary(all_fish_data$percent_large_woody_cover_inchannel)
summary(all_fish_data$percent_submerged_aquatic_veg_inchannel)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$percent_no_cover_inchannel) |> sum()`
NA values There are
`r is.na(all_fish_data$percent_small_woody_cover_inchannel) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_large_woody_cover_inchannel) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_submerged_aquatic_veg_inchannel) |> sum()`
NA values

### Variable: `overhead cover`

**Plotting overhead cover over Period of Record**

Notes: - some cover totals more than 100%

```{r}
all_fish_data |> 
  arrange(river_mile) |> 
  mutate(obs_id = 1:nrow(all_fish_data)) |> 
  select(obs_id, 
         percent_undercut_bank,
         percent_no_cover_overhead,
         percent_cover_half_meter_overhead,
         percent_cover_more_than_half_meter_overhead) |> 
  pivot_longer(cols = percent_undercut_bank:percent_cover_more_than_half_meter_overhead, 
               names_to = "overhead_cover_type", values_to = "percent") |> 
  ggplot(aes(x = obs_id, y = percent, fill = overhead_cover_type)) +
  geom_col() +
  scale_fill_manual(values = colors_small[4:7]) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Numeric Summary of overhead cover over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$percent_undercut_bank)
summary(all_fish_data$percent_no_cover_overhead)
summary(all_fish_data$percent_cover_half_meter_overhead)
summary(all_fish_data$percent_cover_more_than_half_meter_overhead)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$percent_undercut_bank) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_no_cover_overhead) |> sum()` NA values
There are
`r is.na(all_fish_data$percent_cover_half_meter_overhead) |> sum()` NA
values There are
`r is.na(all_fish_data$percent_cover_more_than_half_meter_overhead) |> sum()`
NA values

### Variable: `sur_turb`

**Plotting sur_turb over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = sur_turb)) + 
  geom_histogram(breaks=seq(0, 75, by=1)) + 
  scale_x_continuous(breaks=seq(0, 75, by=5)) +
  theme_minimal() +
  labs(title = "Surface turbidity distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of sur_turb over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$sur_turb)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$sur_turb) |> sum()` NA values

### Variable: `dist_to_bottom`

**Plotting dist_to_bottom over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = dist_to_bottom)) + 
  geom_histogram(breaks=seq(0, 5, by=1)) + 
  scale_x_continuous(breaks=seq(0, 5, by=1)) +
  theme_minimal() +
  labs(title = "Distance") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of dist_to_bottom over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$dist_to_bottom)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$dist_to_bottom) |> sum()` NA values

### Variable: `fish_data_id`

Looks like there are one more unique fish data id than there is number
of rows where joined fish has a count greater than o

```{r}
nrow(all_fish_data |> filter(count > 0)) == length(unique(all_fish_data$fish_data_id))
```

**Numeric Summary of fish_data_id over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$fish_data_id)

```

**NA and Unknown Values**

There are `r is.na(all_fish_data$fish_data_id) |> sum()` NA values

### Variable: `micro_hab_data_tbl_id`

There are more observations than unique micro hab ids so there are some
micro habitat transects that have more than one row in the table

```{r}
nrow(all_fish_data) == length(unique(all_fish_data$micro_hab_data_tbl_id))

```

**Numeric Summary of micro_hab_data_tbl_id over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$micro_hab_data_tbl_id)

```

**NA and Unknown Values**

There are `r is.na(all_fish_data$micro_hab_data_tbl_id) |> sum()` NA
values

### Variable: `water_temp`

**Plotting water_temp over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = water_temp)) + 
  geom_histogram(breaks=seq(0, 90, by=1)) + 
  scale_x_continuous(breaks=seq(0, 90, by=5)) +
  theme_minimal() +
  labs(title = "Temperature distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of water_temp over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$water_temp)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$water_temp) |> sum()` NA values

### Variable: `flow`

**Plotting flow over Period of Record**

```{r}
all_fish_data |> 
  # filter(fork_length < 250) %>% # filter out 13 points so we can more clearly see distribution
  ggplot(aes(x = flow)) + 
  geom_histogram(breaks=seq(0, 3500, by=50)) + 
  scale_x_continuous(breaks=seq(0, 3500, by=1000)) +
  theme_minimal() +
  labs(title = "Flow distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of flow over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$flow)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$flow) |> sum()` NA values

### Variable: `number_of_divers`

**Plotting number_of_divers over Period of Record**

```{r}
all_fish_data |> 
  ggplot(aes(x = number_of_divers)) + 
  geom_histogram(breaks=seq(0, 6, by=1)) + 
  scale_x_continuous(breaks=seq(0, 6, by=1)) +
  theme_minimal() +
  labs(title = "Number of Divers distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of number_of_divers over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$number_of_divers)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$number_of_divers) |> sum()` NA values

### Variable: `reach_length`

All of the reach lengths are 25 (m?)

**Plotting reach_length over Period of Record**

```{r}
all_fish_data |> 
  ggplot(aes(x = reach_length)) + 
  geom_histogram(breaks=seq(0, 100, by=1)) + 
  scale_x_continuous(breaks=seq(0, 100, by=25)) +
  theme_minimal() +
  labs(title = "Reach Length distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of reach_length over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$reach_length)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$reach_length) |> sum()` NA values

### Variable: `reach_width`

All of the reach widths are 4 meters

**Plotting reach_width over Period of Record**

```{r}
all_fish_data |> 
  ggplot(aes(x = reach_width)) + 
  geom_histogram(breaks=seq(0, 10, by=1)) + 
  scale_x_continuous(breaks=seq(0, 10, by=1)) +
  theme_minimal() +
  labs(title = "Reach Width distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of reach_width over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$reach_width)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$reach_width) |> sum()` NA values

### Variable: `chanel_width`

**Plotting chanel_width over Period of Record**

Even through reach width measured is only ever 4 meters sometimes
channel width is much larger

```{r}
all_fish_data |> 
  ggplot(aes(x = channel_width)) + 
  geom_histogram(breaks=seq(0, 170, by=1)) + 
  scale_x_continuous(breaks=seq(0, 170, by=50)) +
  theme_minimal() +
  labs(title = "Channel Width distribution") + 
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**Numeric Summary of chanel_width over Period of Record**

```{r}
# Table with summary statistics
summary(all_fish_data$channel_width)
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$chanel_width) |> sum()` NA values

## Explore Categorical variables: {.tabset}

```{r}
# Filter clean data to show only categorical variables 
all_fish_data %>% select_if(is.character) %>% colnames()
```

### Variable: `location`

```{r}
table(all_fish_data$location) 
```

Fix inconsistencies with spelling, capitalization, and abbreviations.

```{r}
# Fix any inconsistencies with categorical variables
all_fish_data$location <- tolower(all_fish_data$location)
table(all_fish_data$location) 

all_fish_data <- all_fish_data |> 
  mutate(location = case_when(location %in% c("aleck riffle", "alec riffle") ~ "aleck riffle",
                              location == "hatchery  ditch" ~ "hatchery ditch",
                              location %in% c("hour bars", "hour bar") ~ "hour bars",
                              T ~ location))
unique(all_fish_data$location)
# FIX aleck riffle and alec riffle
# FIX hatchery  ditch and hatchery ditch
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$location) |> sum()` NA values

### Variable: `species`

```{r}
table(all_fish_data$species) 
```

Fix inconsistencies with spelling, capitalization, and abbreviations.

**NA and Unknown Values**

There are `r is.na(all_fish_data$species) |> sum()` NA values

### Variable: `channel_geomorphic_unit`

```{r}
table(all_fish_data$channel_geomorphic_unit) 
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$channel_geomorphic_unit) |> sum()` NA
values

### Variable: `gps_coordinate`

```{r}
table(all_fish_data$gps_coordinate) 
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$gps_coordinate) |> sum()` NA values

### Variable: `weather`

```{r}
table(all_fish_data$weather) 
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$weather) |> sum()` NA values

### Variable: `channel_type`

```{r}
table(all_fish_data$channel_type) 
```

**NA and Unknown Values**

There are `r is.na(all_fish_data$channel_type) |> sum()` NA values

## Summary of identified issues

-   percent cover that is \> or \< 100%
-   did some location name clean up but could be better
-   remove sac squawfish
-   other?

## Fixing location names and adding coordinates

The location names were reconciled for the 2001 data (see
`qc_mini_snorkel_data.Rmd`). Here we will read in the reconciled names
and add the 2002 data.

```{r}
microhabitat_with_fish_detections <- all_fish_data |> 
  rename(transect_code = t_code,
         location_table_id = p_dat_id,
         surface_turbidity = sur_turb) |> 
  mutate(species = tolower(species),
         species = ifelse(species == "sacramento squawfish","sacramento pikeminnow", species),
         count = ifelse(is.na(count), 0, count),
         channel_geomorphic_unit = tolower(channel_geomorphic_unit),
         channel_geomorphic_unit = case_when(channel_geomorphic_unit == "glide edgewater" ~ "glide margin",
                                             channel_geomorphic_unit == "riffle edgewater" ~ "riffle margin",
                                             T ~ channel_geomorphic_unit)) |> 
  select(micro_hab_data_tbl_id, location_table_id, transect_code, fish_data_id, date, count, species, fl_mm, dist_to_bottom, depth, focal_velocity, velocity, surface_turbidity, percent_fine_substrate, percent_sand_substrate, percent_small_gravel_substrate, percent_large_gravel_substrate, percent_cobble_substrate, percent_boulder_substrate, percent_no_cover_inchannel, percent_small_woody_cover_inchannel, percent_large_woody_cover_inchannel, percent_submerged_aquatic_veg_inchannel, percent_undercut_bank, percent_no_cover_overhead, percent_cover_half_meter_overhead, percent_cover_more_than_half_meter_overhead, channel_geomorphic_unit) 

survey_locations <- all_fish_data |> 
  rename(transect_code = t_code,
         location_table_id = p_dat_id) |> 
  mutate(weather = tolower(weather),
         channel_type = tolower(channel_type),
         location_revised = case_when(grepl("big hole", location) ~ "big hole",
                                      grepl("auditorium", location) ~ "auditorium riffle",
                                      location %in% c("bedrock", "bedrock park", "bedrock park, unit #50") ~ "bedrock riffle",
                                      grepl("bedrock", location) ~ "bedrock riffle",
                                      grepl("cox", location) ~ "cox riffle", # TODO: is this a new location? 
                                      grepl("gateway", location) ~ "gateway riffle", # TODO: is this a new location? 
                                      grepl("goose", location) ~ "goose riffle", 
                                      grepl("aleck", location) ~ "aleck riffle", 
                                      grepl("big bar", location) ~ "big bar",
                                      grepl("eye", location) ~ "eye riffle",
                                      location %in% c("east g 95", "g 95", "g-95 side channel", "g95", "g95 (area)", "g95 east", "g95 rr downstream head", "g95 west side channel") ~ "g95",
                                      grepl("g95", location) ~ "g95",
                                      grepl("g-95", location) ~ "g95",
                                      grepl("g 95", location) ~ "g95",
                                      grepl("gridley", location)  ~ "gridley riffle",
                                      grepl("grifley", location)  ~ "gridley riffle",
                                      grepl("herringer", location) ~ "herringer riffle",
                                      grepl("junkyard", location) ~ "junkyard riffle",
                                      location == "lower hole" ~ "lower big hole",
                                      location == "lower hour side channel" ~ "lower hour",
                                      grepl("hour", location) ~ "lower hour",
                                      grepl("matthews", location) ~ "matthews riffle",
                                      grepl("mathews", location) ~ "matthews riffle",
                                      grepl("mcfarland", location) ~ "macfarland riffle",
                                      grepl("macfarland", location) ~ "macfarland riffle",
                                      grepl("trailer park", location) ~ "trailer park",
                                      grepl("trialer", location) ~ "trailer park",
                                      grepl("robinson", location) ~ "robinson riffle",
                                      grepl("robinsoon", location) ~ "robinson riffle",
                                      grepl("shallow", location) ~ "shallow riffle",
                                      grepl("steep", location) ~ "steep riffle",
                                      grepl("vance", location) ~ "vance avenue",
                                      grepl("weir", location) ~ "weir riffle",
                                      grepl("hatchery riffle", location) ~ "hatchery riffle", 
                                      grepl("hatchery ditch", location) ~ "hatchery ditch", 
                                      T ~ location)) |> 
  select(location_table_id, date, location, location_revised, water_temp, 
         weather, river_mile, flow, number_of_divers, 
         reach_length, reach_width, channel_width, 
         channel_type, gps_coordinate) |> 
  select(-location) |> 
  rename(location = location_revised) |> 
  distinct() 

```

Okay, let's compare the names of the 2001/2002 data with the reconciled
name list from 2001.

```{r}
# This is created in the qc_mini_snorkel_data.Rmd
survey_locations_2001 <- read_csv(here::here("edi-mini-snorkel", "data", "archive", "survey_locations.csv")) 

# the only two names that exist in the 2002 dataset are and NOT the 2001 are: cox riffle and gateway riffle
setdiff(sort(unique(survey_locations$location)), sort(unique(survey_locations_2001$location)))

# create a location lookup using the 2001 reconciled locations that can be used for 2002 data 
lookup <- survey_locations_2001 |> 
  select(location, channel_type, longitude, latitude, coordinate_method, channel_location) # river_mile

# joined by location and channel type
location_data_updated <- survey_locations |> 
  filter(year(date) == 2002) |> 
  left_join(lookup) |> 
  select(colnames(survey_locations_2001)) |> 
  bind_rows(survey_locations_2001) |> 
  arrange(location_table_id) |> 
  mutate(latitude = case_when(location == "cox riffle" ~  39.33564114789087, # based off of gps_coordinates
                              location == "gateway riffle" ~ 39.45696577928121, 
                              (location == "shallow riffle" & channel_type == "mainchannel branch") ~ 39.32967570030937,
                              (location == "shallow riffle" & channel_type == "mainchannel") ~ 39.33154115157018, 
                              T ~ latitude), 
         longitude = case_when(location == "cox riffle" ~ -121.63215000396984, # based off of gps_coordinates
                               location == "gateway riffle" ~ -121.62560364814165,
                               (location == "shallow riffle" & channel_type == "mainchannel branch") ~ -121.6285435907359,
                               (location == "shallow riffle" & channel_type == "mainchannel") ~ -121.63033927513442,
                               T ~ longitude), 
         coordinate_method = case_when(location %in% c("cox riffle", "gateway riffle") ~ "assigned based on similar location",
                                       (location == "shallow riffle" & channel_type %in%  c("mainchannel branch", "mainchannel")) ~ "assigned based on similar location", 
                                       T ~ coordinate_method),
         channel_location = case_when(location == "shallow riffle" ~ "HFC",
                                      location == "trailer park" ~ "LFC", # this removes 1 NA but is consistent with all other trailer parks            
                                      location == "cox riffle" ~ "HFC",
                                      location == "gateway riffle" ~ "LFC", 
                                      T ~ channel_location),
         channel_type = case_when(location == "trailer park" ~ "mainchannel",
                                  T ~ channel_type)) # this removes 1 NA but is consistent with all other trailer parks

# QC
# test_2001 <- location_data_2002 |> 
#   filter(year(date) == 2001) 
# unique(test_2001 == survey_locations_2001)

```


Save data:

```{r}

write_csv(location_data_updated, here::here("data", "survey_locations.csv"))
write_csv(microhabitat_with_fish_detections, here::here("data", "microhabitat_observations.csv"))

```

```{r}

knitr::knit_exit()

```
