
```{r, include=FALSE}
library(tidyverse)
library(googleCloudStorageR)
library(ggplot2)
```

```{r, include=FALSE}
# google cloud set up
gcs_auth(json_file = Sys.getenv("GCS_AUTH_FILE"))
gcs_global_bucket(bucket = Sys.getenv("GCS_DEFAULT_BUCKET"))
```

```{r, include=FALSE}
# get data from google cloud
gcs_get_object(object_name = "juvenile-rearing-monitoring/seine-and-snorkel-data/feather-river/data/combined_feather_snorkel_data.csv",
               bucket = gcs_get_global_bucket(),
               saveToDisk =  here::here("data-raw", "combined_feather_snorkel_data.csv"),
               overwrite = TRUE)
combined_snorkel <- read_csv(here::here("data-raw", "combined_feather_snorkel_data.csv"))
```


## Fields that still need clarification from Casey or team:
 - section_name
 - location
 - hydrology
 - unit_type
 

## Comparing section_name and location
These two variables have similar descriptions

## section_name

```{r, echo=FALSE}
table(combined_snorkel$section_name)
```


### Plot showing section_name distribution

```{r, echo=FALSE}
ggplot(combined_snorkel, aes(x = section_name)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
```


### location
```{r, echo=FALSE}
table(combined_snorkel$location)
```

```{r,include=FALSE}
#hard to read distributioinn of location
ggplot(combined_snorkel, aes(x = location)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
```


### Plot showing location distribution

Notice that location has many more fields than section_name
```{r,echo=FALSE}
#plot with NAs removed, still hard to read
ggplot(combined_snorkel[!is.na(combined_snorkel$location),], aes(x = location)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
```


### Names that are consistent between section_name and location:

```{r, echo=FALSE}
intersect(combined_snorkel$section_name, combined_snorkel$location)
```

### Fields that are different between section_name and location

```{r, echo=FALSE}
setdiff(combined_snorkel$section_name, combined_snorkel$location)
```

## Comparing unit_type and hydrology
 
 
### unit_type
```{r, echo=FALSE}
table(combined_snorkel$unit_type)
```


### Plot showing unit_type distribution
```{r, echo=FALSE}
ggplot(combined_snorkel, aes(x = unit_type)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
```



### hydrology
```{r, echo=FALSE}
table(combined_snorkel$hydrology)
```

### Plot showing hydrology distribution
```{r, echo=FALSE}
ggplot(combined_snorkel, aes(x = hydrology)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
```


### Plot showing overlap between hydrology and unit_type
```{r, echo=FALSE}
ggplot(combined_snorkel, aes(x = hydrology, fill = "hydrology")) +
  geom_bar(position = position_dodge(width = 0.9)) +
  geom_bar(aes(x = unit_type, fill = "unit_type"), position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("hydrology" = "red", "unit_type" = "black"))
```


### Names that are consistent between unit_tyoe and hydrology:
```{r, echo=FALSE}
intersect(combined_snorkel$unit_type, combined_snorkel$hydrology)
```


### Fields that are different between unit_tyoe and hydrology
```{r, echo=FALSE}
setdiff(combined_snorkel$section_name, combined_snorkel$location)
```


