---
title: "Feather River - Mini Snorkel Data Modeling and Analysis for Steelhead"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3      
    toc_float: true   
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(sjPlot)
library(pROC)
library(glmmTMB)
library(sf)
library(leaflet)
library(scales)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 11, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    )
)


custom_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

colors_full <-  c("#9A8822", "#F5CDB4", "#F8AFA8", "#FDDDA0", "#74A089", #Royal 2
                  "#899DA4", "#C93312", "#DC863B", # royal 1 (- 3)
                  "#F1BB7B", "#FD6467", "#5B1A18", "#D67236",# Grand Budapest 1 (-4)
                  "#D8B70A", "#02401B", "#A2A475", # Cavalcanti 1
                  "#E6A0C4", "#C6CDF7", "#D8A499", "#7294D4", #Grand Budapest 2
                  "#9986A5", "#EAD3BF", "#AA9486", "#B6854D", "#798E87", # Isle of dogs 2 altered slightly
                  "#F3DF6C", "#CEAB07", "#D5D5D3", "#24281A", # Moonriese 1, 
                  "#798E87", "#C27D38", "#CCC591", "#29211F", # moonrise 2
                  "#85D4E3", "#F4B5BD", "#9C964A", "#CDC08C", "#FAD77B" ) # moonrise 3 

options(scipen=999)

```

## Objective

Develop a model that reflects the significance of cover, substrate,
depth, and velocity on steelhead presence and absence.

## Analysis Overview

Model was developed using [Feather River Mini Snorkel
Data](https://github.com/Healthy-Rivers-and-Landscapes-Science/feather-river/tree/main/edi-mini-snorkel)
filtered to steelhead observations. The dataset consists of numeric fish
count measurements that can also be represented as a binary
presence–absence response. Because steelhead counts were highly
zero‐inflated, with many transects recording no fish and a small number
of observations containing relatively large counts, we initially
explored abundance‐based modeling approaches (ie., Hurdle models).
However, preliminary analyses indicated that models attempting to
predict steelhead counts performed very poorly, reflecting the high
variability and limited information content in the abundance data. As a
result, abundance modeling was not pursued further and is not included
in this document.

Subsequent analyses therefore focused on modeling steelhead occurrence
using logistic regression, which better matched the structure and signal
contained in the data. Mini Snorkel surveys were conducted between March
and August, a period that spans substantial seasonal changes in habitat
conditions and steelhead outmigration, resulting in strong temporal
variation in presence. Steelhead were observed predominantly in low-flow
channels, reflecting differences in habitat availability, hydraulic
conditions, and refuge opportunities between channel types. Although
initial exploratory analyses considered fitting separate models for
high-flow and low-flow channels, we ultimately adopted a unified
logistic regression framework that incorporated spatial and temporal
heterogeneity through random effects. This approach allowed us to
account for systematic differences among sites and seasons while
retaining a single, interpretable model structure for inference.

To account for spatial and temporal structure, we evaluated a series of
mixed‐effects logistic regression models that included random intercepts
for transect site and month. These random effects captured persistent
differences among sites, repeated sampling within transects, and
seasonal shifts in steelhead presence. Among the candidate models, the
best‐performing model included random effects for both site and month,
highlighting the importance of spatial heterogeneity and seasonal
dynamics in shaping steelhead occurrence. Including channel type as an
additional random effect did not improve model performance. This final
model provided the strongest overall fit and formed the basis for
inference on habitat associations with steelhead presence.

## Review Data

```{r}
source(here::here("mini-snorkel-analysis", "data-raw", "pull_from_edi.R"))

mini_snorkel_raw <- mini_fish_raw |> 
  left_join(mini_locations_raw |> distinct()) |> 
  mutate(count = ifelse(is.na(count), 0, count),
         fish_presence = as.factor(ifelse(count < 1, "0", "1")),
         month = month(date)) |> 
  filter(species %in% c("steelhead trout (wild)", "steelhead trout (clipped)") | count == 0) |> 
  glimpse()

```

## Data Exploration

### High Flow vs. Low Flow Channel

Most of the steelhead were found in the Low Flow channel.

```{r echo=FALSE, message=FALSE, warning=FALSE}
mini_snorkel_raw |> 
  filter(count > 0) |> 
  ggplot() + 
  geom_histogram(aes(count, fill = channel_location), binwidth = 50) + 
  scale_fill_manual(values = custom_colors) + 
  ylab("steelhead count") +
  ggtitle('Figure 1. Histogram of steelhead count in the high and low flow channels') +facet_grid(~channel_location)

mini_snorkel_model_ready <- mini_snorkel_raw
```

```{r}

# looking at overall count data between the low flow/high flow channels
mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  tally(count) |> 
  knitr::kable(caption = "Table 1. Total count of chinook salmon between high flow and low flow channels")

mini_snorkel_model_ready |> 
  group_by(channel_location) |> 
  summarize(n_sites = length(unique(location))) |> 
  knitr::kable(caption = "Table 2. Number of sampling sites in the high flow and low flow channels")
```

### Temporal exploration

The majority of fish are found in the low flow channel with the highest
counts in April.

```{r echo=FALSE, message=FALSE, warning=FALSE}

mini_snorkel_model_ready |> 
  group_by(channel_location, month = lubridate::month(date)) |> 
  tally(count) |> 
  ggplot() + 
  geom_col(aes(x = as.factor(month), y = n, fill = channel_location), position = "dodge") + 
  ylab("fish count") +
  xlab("month") +
  ggtitle('Figure 2. Steelhead presence as a function of month and high flow/low flow channel') + 
  scale_fill_manual(values = custom_colors)

```

### Redd Data Integration and Exploration

**Data Question:** - for sites that include roving and transect, can I
aggregate or is that double counting? Currently grouped by date and
location name to get a total number of redds.

```{r}
redd_data_raw <- readxl::read_excel(here::here("mini-snorkel-analysis", "data-raw", "SH Redd Survey.xlsx")) |> 
  janitor::clean_names() 

# remove zero redds
redd_data_with_redds <- redd_data_raw |> 
  rename(number_redds = number_of_redds) |> 
  filter(number_redds > 0) |> 
  mutate(lat = as.numeric(lat),
         long = as.numeric(long)) |>  
  filter(!is.na(lat) | !is.na(long))
```

**Redd data over time**

This visual represents the number of redds over time at each location.
It helps provide context on which sites generally have redds and if they
have redds consistently over time.

```{r}
redd_data_with_redds |>
  group_by(date, location, lat, long) |>
  summarise(total_redds = sum(number_redds), .groups = "drop") |>
  group_by(location) |>
  summarise(
    lat0 = median(lat, na.rm = TRUE),
    .groups = "drop"
  ) |>
  right_join(
    redd_data_with_redds |>
      group_by(date, location) |>
      summarise(total_redds = sum(number_redds), .groups = "drop"),
    by = "location"
  ) |>
  mutate(
    location = fct_reorder(location, lat0, .desc = TRUE),  # north at top
    year = year(date)
  ) |>
  ggplot(aes(x = year, y = location, fill = total_redds)) +
  geom_tile(color = "white", linewidth = 0.5) +
  xlab("Year") +
  ylab("Location (ordered South to North)")
```

```{r eval=FALSE, include=FALSE}
# QC map for locations North/South
redd_data_by_location <- redd_data_with_redds |> 
  rename(date_redd = date) |> 
  filter(!is.na(lat)) |> 
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326) 

# steelhead_raw_redds <- redd_data_by_location |> 
#   group_by(location) |> 
#   summarise(total_redds = sum(number_redds, na.rm = T)) |> 
#   saveRDS(here::here("mini-snorkel-analysis", "visualizations", "raw_redd_location_steelhead.rds"))

leaflet() |> 
  addProviderTiles(providers$Esri.WorldImagery, group = "Aerial Imagery") |>
  addProviderTiles(providers$OpenStreetMap, group = "Street Map") |>
  addCircleMarkers(data = redd_data_by_location, 
                     layerId = ~location, 
                     color = "darkblue",
                     radius = 6,
                     fill = TRUE,
                     fillOpacity = 0.6,
                     opacity = 0.8,
                     popup = paste0("Location Name: ", redd_data_by_location$location) )
  
```

Total number of redds by year

```{r}
redd_by_year <- redd_data_with_redds |> group_by(year = year(date)) |> summarise(total_redds = sum(number_redds)) 

knitr::kable(redd_by_year)

```

Combine redd data with mini snorkel to see if redds are a spatial
indicator or spawning potential/habitat quality.

```{r}
redd_data_by_location <- redd_data_with_redds |> 
  rename(date_redd = date) |> 
  filter(!is.na(lat)) |> 
  sf::st_as_sf(coords = c("long", "lat"), crs = 4326)

# redd_data_by_location |> 
#   ggplot(aes(x = location, y = n_redds)) +
#   geom_col() +
#   coord_flip()

mini_sf <- mini_snorkel_model_ready |> 
  filter(!is.na(longitude)) |> 
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 
```

Find the nearest transect location from the Mini Snorkel data to each
redd

```{r}
dist_to_nearest <- st_distance(
  redd_data_by_location,
  mini_sf[st_nearest_feature(redd_data_by_location, mini_sf), ],
  by_element = TRUE
)

redd_near <- redd_data_by_location |>
  mutate(dist_to_nearest = as.numeric(dist_to_nearest)) |> 
  filter(dist_to_nearest <= 50)

redd_near_with_snorkel <- st_join(
  redd_near |> select(-location),
  mini_sf,
  join = st_nearest_feature
)
```

Redds are joined to the nearest mini snorkel transect location within
50-meters. The following histogram shows the distribution of the redd
distances; 50-meters was chosen based off of the high counts (shown as
the red dashed line). Note that three outliers were removed, shown in
the table below. The locations (which are visible as light blue circles
in the leaflet map) look to be errors.

```{r}
redd_outliers <- redd_data_by_location |>
  mutate(dist_to_nearest = as.numeric(dist_to_nearest)) |> 
  filter(dist_to_nearest > 1000)

knitr::kable(redd_outliers |> st_drop_geometry() |> select(date = date_redd, location, number_redds), caption = "Outliers - The following three data points seem to have incorrect coordinates")

redd_data_by_location |>
  mutate(dist_to_nearest = as.numeric(dist_to_nearest)) |> 
  filter(dist_to_nearest < 1000) |> # removing outliers
  ggplot(aes(x = dist_to_nearest)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  geom_vline(xintercept = 50, linetype = 2, color = "red") +
  labs(
    title = "Distribution of nearest-neighbor distances"
  ) +
  theme_minimal(base_size = 13)

redd_summary <- redd_near_with_snorkel |> 
  st_drop_geometry() |> 
  group_by(location) |> 
  summarise(
    redd_total = sum(number_redds),
    redd_mean = mean(number_redds),
    redd_median = median(number_redds),
    redd_presence = as.integer(redd_total > 0)
  )
```

A visual representation of the amount of redds joined with each of the
mini snorkel transect locations. Outliers that have been removed are
shown in blue.

```{r}

locations_sf <- sf::st_as_sf(mini_locations_raw |> 
                               distinct(location_table_id, date, channel_location, .keep_all = T) |> 
                               na.omit(), 
                             coords = c('longitude', 'latitude')) |> 
  select(location_table_id, location, channel_type, date) |> 
  mutate(location_table_id = paste0(date, " (location id: ", location_table_id, ")")) |> 
  mutate(color = colors_full[match(location, location)]) |> 
  group_by(location) |> 
  slice(1) |> # This is a bit of a workaround since there are multiple lat/longs for reach location that are near eachother. Since this is just for a visual, I think it's fine. 
  ungroup()

redd_summary_sf <- redd_summary |> 
  left_join(locations_sf) |> 
  select(location, redd_presence, redd_total, geometry) |> 
  distinct() |> 
  sf::st_as_sf(sf_column_name = "geometry", crs = 4326)

leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Aerial Imagery") |>
  addProviderTiles(providers$OpenStreetMap, group = "Street Map") |>
  addCircleMarkers(data = redd_outliers,
                   popup = paste0("Outlier Location Name: ", redd_outliers$location,"<br>",
                                  "Date: ", redd_outliers$date_redd)) |> 
   addCircleMarkers(data = locations_sf, 
                     layerId = ~location, 
                     color = ~color,
                     radius = 6,
                     fill = TRUE,
                     fillOpacity = 0.2,
                     opacity = 0.6,
                     popup = paste0("Location Name: ", locations_sf$location,
                                    "<br>",
                                    "Channel Type: ", locations_sf$channel_type)) |> 
  addCircleMarkers(data = redd_summary_sf, 
                   color = "darkred",
                   radius = ~rescale(redd_total, to = c(3, 20)),
                   fill = TRUE,
                   fillOpacity = 1,
                   opacity = 1, 
                   popup = paste0("Location Name: ", redd_summary_sf$location,
                                    "<br>",
                                    "N Redds: ", redd_summary_sf$redd_total)) |> 
   
  addLayersControl(
    baseGroups = c("Aerial Imagery", "Street Map"),
    position = "topleft",
    options = layersControlOptions(collapsed = FALSE)
  ) |> 
  addLegend(colors = "darkred", labels = "Redd Locations") 

```

### Variables of Interest

The variables of interest include cover, substrate, velocity and depth
variables known to be important for steelhead rearing habitat. We are
also including the number of redds found near each mini snorkel transect
and whether or not there were redds nearby.

-   Velocity - numeric
-   Depth - numeric
-   Number of Redds - numeric
-   Redd at location - 0/1
-   Month - categorical (3-8)

Substrate and cover variables are measured as percentages are converted
to binary presence/absence (1/0) by establishing a threshold percentage
of 20%. Overhanging vegetation was measured at 1/2 meter overhead and
more than 1/2 meter overhead. These categories were combined for
simplicity and for comparison with other studies, such as those by Mark
Gard.

-   Undercut bank - 0/1
-   Aquatic vegetation - 0/1
-   Overhanging vegetation - 0/1
-   Small woody cover - 0/1
-   Large woody cover - 0/1
-   Boulder substrate - 0/1
-   Cobble substrate - 0/1

## Build Model Data

All cover variables were converted to presence/absence using a threshold
of 20%. The following is the data structure of the model input data.

```{r}

percent_threshold <- 20

# use a hurdle model (Gard 2024); 
# pre-processing - all numeric values must be rounded to whole numbers
model_data <- mini_snorkel_model_ready |>
  select(count, location, channel_location, depth, velocity, contains("inchannel"), contains("overhead"), "percent_cobble_substrate", "percent_boulder_substrate", "percent_undercut_bank", month, fl_mm, channel_geomorphic_unit, reach_length, reach_width, channel_type, month, surface_turbidity) |>
  # create new cover variables, based off Gard 2023
  mutate(small_woody = percent_small_woody_cover_inchannel,
         large_woody = percent_large_woody_cover_inchannel,
         boulder_substrate = percent_boulder_substrate,
         cobble_substrate = percent_cobble_substrate, 
         undercut_bank = percent_undercut_bank, 
         aquatic_veg = percent_submerged_aquatic_veg_inchannel, 
         overhanging_veg = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |>
  # transform to presence/absence based on a defined threshold
  mutate(cobble_substrate = ifelse(percent_cobble_substrate >= percent_threshold, 1, 0 ),
         boulder_substrate = ifelse(percent_boulder_substrate >= percent_threshold, 1, 0 ),
         small_woody = ifelse(small_woody >= percent_threshold, 1, 0 ),
         large_woody = ifelse(large_woody >= percent_threshold, 1, 0 ),
         aquatic_veg = ifelse(aquatic_veg >= percent_threshold, 1, 0),
         undercut_bank = ifelse(undercut_bank >= percent_threshold, 1, 0),
         no_cover_overhead = ifelse(percent_no_cover_overhead >= percent_threshold, 1, 0),
         overhanging_veg = ifelse(overhanging_veg >= percent_threshold, 1, 0)) |>
  #remove no cover variables (no cover overhead, no cover in channel)
  select(-contains("no_cover")) |> 
  distinct() |> 
  mutate(fl_mm = ifelse(is.na(fl_mm), 0, fl_mm)) |> 
  na.omit() |> 
  select(-fl_mm) |> 
  left_join(redd_summary) |> 
  mutate(
    redd_total = replace_na(redd_total, 0),
    redd_presence = replace_na(redd_presence, 0),
    month = as.factor(month), 
  ) |> 
  glimpse()

```

## Model Performance Evaluation Overview

Model performance was evaluated using a combination of receiver
operating characteristic (ROC) analysis and confusion matrix–based
classification metrics. The area under the ROC curve (AUC) was used to
assess the model’s ability to discriminate between presence and absence
across all possible probability thresholds. AUC provides a
threshold-independent measure of performance, where values near 0.5
indicate no discrimination and higher values indicate increasing ability
to correctly rank presences above absences.

In addition, model predictions were converted to binary classifications
using a fixed probability threshold, and a confusion matrix was used to
summarize agreement between predicted and observed outcomes. The
confusion matrix tabulates true positives, true negatives, false
positives, and false negatives, allowing evaluation of classification
behavior such as sensitivity to presences and specificity to absences.
Because the dataset was imbalanced, with many more absences than
presences, emphasis was placed on interpreting the confusion matrix in
conjunction with AUC rather than relying on overall accuracy alone.

Threshold predictive probabilities into 0/1 set at `0.5`.

#### Models were evaluated by the following metrics:

-   **N** Total number of observations used to evaluate the model.

-   **True Positives (TP)** Cases where the model predicted **presence
    (1)** and fish were actually **present (1)**. *Correct detections.*

-   **False Positives (FP)** Cases where the model predicted **presence
    (1)** but fish were actually **absent (0)**. *False alarms.*

-   **True Negatives (TN)** Cases where the model predicted **absence
    (0)** and fish were actually **absent (0)**. *Correct
    non-detections.*

-   **False Negatives (FN)** Cases where the model predicted **absence
    (0)** but fish were actually **present (1)**. *Missed detections.*

-   **Accuracy** The overall proportion of correct predictions: (TP +
    TN) / N Note: Accuracy can appear high when absences are much more
    common than presences.

-   **Sensitivity (Recall)** Of all true presences, the proportion
    correctly detected: TP / (TP + FN) Measures how well the model
    detects fish when they are present.

-   **Specificity** Of all true absences, the proportion correctly
    predicted as absent: TN / (TN + FP) Measures how well the model
    avoids false alarms.

-   **Precision (Positive Predictive Value, PPV)** Of all predicted
    presences, the proportion that are truly present: TP / (TP + FP)
    Indicates how trustworthy a “presence” prediction is.

-   **Negative Predictive Value (NPV)** Of all predicted absences, the
    proportion that are truly absent: TN / (TN + FN) Indicates how
    trustworthy an “absence” prediction is.

-   **Balanced Accuracy** The average of sensitivity and specificity:
    (Sensitivity + Specificity) / 2 More informative than overall
    accuracy when classes are imbalanced.

-   **Prevalence** The proportion of observations where fish were
    present: (TP + FN) / N Represents how common fish presence is in the
    dataset.

-   **Area Under the Curve (AUC)**AUC refers to the *Area Under the
    Receiver Operating Characteristic (ROC) Curve*. It measures how well
    a model can distinguish between presence (1) and absence (0) across
    all possible probability thresholds, not just one cutoff.

```{r}
threshold <- 0.5
```

## Logistic Regression Approach

Because fish counts were highly zero-inflated and exhibited extreme
variability among non-zero values, abundance models performed poorly. We
therefore focused on modeling fish occurrence using logistic regression,
which better matches the information content of the data and provides
more reliable inference on habitat associations.

```{r}
log_reg_data <- model_data |> 
  mutate(presence = as.integer(count > 0))
```

### Build Logistic Regression Model

A simple logistic regression to start.

```{r}
glm_presence <- glm(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + surface_turbidity,
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glm_presence)

```

```{r}

pred_prob <- predict(glm_presence, type = "response")
obs <- log_reg_data$presence

roc_glm <- roc(obs, pred_prob)
auc(roc_glm)   # slightly better performing 
```

AUC of `r auc(roc_glm)` means The habitat variables explain some
presence–absence structure, but there is still a lot of overlap/noise in
the data.

Confusion Matrix:

```{r echo=FALSE}
# confusion matrix 
pred_class <- ifelse(pred_prob > threshold, 1, 0)
con_log <- table(Predicted = pred_class, Observed = obs)
con_log
```

### Logistic Regression with Random Effect

### Random Effect of Location

Using the high flow, low flow channel as the random effect:

```{r}

glmm_presence_location <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + surface_turbidity + 
    (1 | location),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence_location)

pred_prob_glmm_loc <- predict(glmm_presence_location, type = "response")
obs <- log_reg_data$presence

roc_glmm_loc <- roc(obs, pred_prob_glmm_loc)
auc(roc_glmm_loc) 

# confusion matrix 
pred_class_loc <- ifelse(pred_prob_glmm_loc > threshold, 1, 0)
con_glmm_loc <- table(Predicted = pred_class_loc, Observed = obs)
con_glmm_loc 
```

### Random Effect of Channel Location \| Location

Nesting the transect location within the high flow/low flow channel as
the random effect:

```{r}

glmm_presence <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + surface_turbidity + 
    (1 | channel_location/location),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence)

```

```{r}
pred_prob_glmm <- predict(glmm_presence, type = "response")
obs <- log_reg_data$presence

roc_glmm <- roc(obs, pred_prob_glmm)
auc(roc_glmm) 

```

```{r echo=FALSE}
# confusion matrix 
pred_class <- ifelse(pred_prob_glmm > threshold, 1, 0)
con_glmm_presence <- table(Predicted = pred_class, Observed = obs)
con_glmm_presence
```

### Random Effect of Location and Month

Using location and month as the random effect.

Based off of the data exploration with majority of steelhead in the low
flow channel, it is unsurprising that location is a better predictor.

```{r}

glmm_presence_location_month <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + surface_turbidity + 
    (1 | location) + (1 | month),                     
  data = log_reg_data,
  family = binomial(link = "logit")
)

summary(glmm_presence_location_month)

pred_prob_glmm_loc_month <- predict(glmm_presence_location_month, type = "response")
obs <- log_reg_data$presence

roc_glmm_loc_month <- roc(obs, pred_prob_glmm_loc_month)
auc(roc_glmm_loc_month) 

# confusion matrix 
pred_class_loc_month <- ifelse(pred_prob_glmm_loc_month > threshold, 1, 0)
con_glmm_presence_location_month <- table(Predicted = pred_class_loc_month, Observed = obs)
con_glmm_presence_location_month
```

### Random Effect of Location and Month (March - May)

Using location and month (subset to march - may) as the random effect.

The main difference in the results in the inclusion of depth as an
important predictor. The AUC is the same as the model prior (with all
months).

```{r}
log_data_month_subset <- log_reg_data |> 
  filter(month %in% c('3', '4', '5', '6')) |> 
  glimpse()

# only signficant predictor difference is depth is significant whereas
# it was not in the all-month model
glmm_presence_location_month_subset <- glmmTMB(
  presence ~ small_woody + depth + velocity + large_woody +
    aquatic_veg + overhanging_veg + cobble_substrate +
    boulder_substrate + undercut_bank + redd_total + redd_presence + surface_turbidity + 
    (1 | location) + (1 | month),                     
  data = log_data_month_subset,
  family = binomial(link = "logit")
)

summary(glmm_presence_location_month_subset)

pred_prob_glmm_loc_month_subset <- predict(glmm_presence_location_month_subset, type = "response")
obs <- log_data_month_subset$presence

roc_glmm_loc_month_subset <- roc(obs, pred_prob_glmm_loc_month_subset)
auc(roc_glmm_loc_month_subset) 

# confusion matrix 
pred_class_loc_month_subset <- ifelse(pred_prob_glmm_loc_month_subset > threshold, 1, 0)
con_glmm_presence_location_month_subset <- table(Predicted = pred_class_loc_month_subset, Observed = obs)
con_glmm_presence_location_month_subset
```

### Model Summary Table

All models showed high overall accuracy (96–97%), largely driven by the
low prevalence of fish presence (3–5%). However, simple logistic and
spatial-only random effects models had very low sensitivity (6–10%),
indicating that they missed most true presences despite good
specificity. Adding both location and month random effects substantially
improved detection (sensitivity 20–27%) while maintaining high precision
and excellent discrimination (AUC ≈ 0.91). The model restricted to
biologically relevant months (3–6) performed best overall, achieving the
highest sensitivity and balanced accuracy while retaining strong AUC.

```{r include=FALSE}
metrics_from_cm <- function(cm, model_name, auc) {

  TN <- cm["0","0"]
  FN <- cm["0","1"]
  FP <- cm["1","0"]
  TP <- cm["1","1"]

  n <- TN + FN + FP + TP

  tibble(
    model = model_name,
    n = n,
    TP = TP,
    FP = FP,
    TN = TN,
    FN = FN,

    accuracy  = (TP + TN) / n,
    sensitivity = TP / (TP + FN),      # the percent of true presences
    specificity = TN / (TN + FP), # percent at identifying absences 
    precision   = TP / (TP + FP),
    npv         = TN / (TN + FN),
    balanced_accuracy = ((TP / (TP + FN)) + (TN / (TN + FP))) / 2,
    prevalence = (TP + FN) / n
  ) |> 
    mutate(auc = auc)
}

```

```{r}
metrics_from_cm(con_log, "simple logistic regression", round(auc(roc_glm), 3)) |> 
  bind_rows(metrics_from_cm(con_glmm_loc, "random effect of location", round(auc(roc_glmm_loc), 3))) |> 
  bind_rows(metrics_from_cm(con_glmm_presence, "random effect of channel_location | location", round(auc(roc_glmm), 3 ))) |> 
  bind_rows(metrics_from_cm(con_glmm_presence_location_month, "random effect of location and month", round(auc(roc_glmm_loc_month), 3 ))) |> 
  bind_rows(metrics_from_cm(con_glmm_presence_location_month_subset, "random effect of location and months (3-6)", round(auc(roc_glmm_loc_month_subset), 3))) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  rename(
    `Model` = model,
    `N` = n,
    `True Positives` = TP,
    `False Positives` = FP,
    `True Negatives` = TN,
    `False Negatives` = FN,
    `Accuracy` = accuracy,
    `Sensitivity (Recall)` = sensitivity,
    `Specificity` = specificity,
    `Precision (PPV)` = precision,
    `NPV` = npv,
    `Balanced Accuracy` = balanced_accuracy,
    `Prevalence` = prevalence,
    `AUC` = auc
  ) |> knitr::kable()

```

### Figures

The following figures are for the model using location and months 3-6 as
the random effect.

The following figure represents the overall effect of month on fish
presence while keeping all other months constant. Points greater than 1
represent a higher probability of fish presence and points less than 1
represent a lower probability. Wider bars represent greater uncertainty.

```{r}

month_effects_df <-  broom.mixed::tidy(glmm_presence_location_month_subset, effects = "ran_vals", conf.int = TRUE) |>
  filter( group == "month") |> 
  rename(unit = level, re = estimate) |>
  mutate(
    odds_ratio = exp(re),
    or_low  = exp(conf.low),
    or_high = exp(conf.high)
  ) |>
  arrange(desc(unit)) |>
  distinct() |>
  mutate(unit = factor(unit, levels = unit)) 

ggplot(month_effects_df, aes(x = odds_ratio, y = unit)) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  labs(
    x = "Random-effect odds ratio (relative to global mean)",
    y = NULL,
    title = "Month effects on probability of presence of steelhead"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 13)

```

The following figure represents the overall effect of each of the sites
on fish presence while keeping all other sites constant. Points greater
than 1 represent a higher probability of fish presence and points less
than 1 represent a lower probability. Wider bars represent greater
uncertainty.

Larger impact for low flow channel.

```{r}
re_loc <- broom.mixed::tidy(glmm_presence_location_month_subset, effects = "ran_vals", conf.int = TRUE) |>
  filter(group == "location") |>
  rename(unit = level, re = estimate) |>
  left_join(mini_snorkel_raw |> select(unit = location, channel_location),
            by = "unit") |>
  mutate(
    odds_ratio = exp(re),
    or_low  = exp(conf.low),
    or_high = exp(conf.high)
  ) |>
  arrange(odds_ratio) |>
  distinct() |>
  mutate(unit = factor(unit, levels = unit)) 

ggplot(re_loc, aes(x = odds_ratio, y = unit, color = channel_location)) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  labs(
    x = "Random-effect odds ratio (relative to global mean)",
    y = NULL,
    title = "Site effects on probability of presence"
  ) +
  scale_color_manual(
    values = c(
      "HFC" = "#1f78b4",   # blue
      "LFC" = "#33a02c"    # green
    ),
    labels = c(
      "HFC" = "High-flow channel",
      "LFC" = "Low-flow channel"
    ),
    name = "Channel type"
  ) +
  scale_x_log10() +
  theme_minimal(base_size = 13)




```

The following figure represents the overall effect of each of the
predictors for fish presence when keeping all other predictors constant.
Points greater than 1 represent a higher probability of fish presence
and points less than 1 represent a lower probability. Wider bars
represent greater uncertainty. The magnitude of these effects is modest
on a per-unit basis, but their significance reflects the precision of
the estimates and the large sample size rather than a lack of biological
relevance.

```{r eval=FALSE, include=FALSE}
# previous plot at log-scale x axis. It was requested to not scale the axis so that 0 is present. 
effects_df <- broom::tidy(glmm_presence_location_month_subset, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  filter(term != "sd__(Intercept)") |> 
  mutate(
    odds_ratio = exp(estimate),
    or_low = exp(conf.low),
    or_high = exp(conf.high),
    term = recode(term,
                  small_woody = "Small woody debris*",
                  large_woody = "Large woody debris",
                  overhanging_veg = "Overhanging vegetation*",
                  aquatic_veg = "Aquatic vegetation*",
                  cobble_substrate = "Cobble substrate",
                  boulder_substrate = "Boulder substrate",
                  undercut_bank = "Undercut bank",
                  depth = "Depth*",
                  velocity = "Velocity*",
                  redd_total = "Total Number or Redd*",
                  redd_presence = "Redds Nearby*",
                  surface_turbidity = "Surface Turbulence"
    )
  )

ggplot(effects_df, aes(x = odds_ratio, y = reorder(term, odds_ratio))) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray50") +
  geom_point(size = 2.7) +
  geom_errorbarh(aes(xmin = or_low, xmax = or_high), height = 0.2) +
  scale_x_log10() +
  labs(
    x = "Odds ratio (log scale)",
    y = NULL,
    title = "Effect sizes for predictors of steelhead presence", 
    caption = "* represents model significance p<0.05"
  ) +
  theme_minimal(base_size = 13)

```

```{r}
# This plot is updated to not have a log-scale for the x-axis. 
# Odds ratio are inherently on the log-scale so the label is still log odds ratio


effects_df <- broom::tidy(glmm_presence_location_month_subset, conf.int = TRUE) |>
  filter(!term %in% c("(Intercept)", "sd__(Intercept)")) |>
  mutate(
    odds_ratio = exp(estimate),
    or_low  = exp(conf.low),
    or_high = exp(conf.high),

    log_or      = estimate,   # log(OR)
    log_or_low  = conf.low,
    log_or_high = conf.high,

    term = recode(term,
      small_woody       = "Small woody debris*",
      large_woody       = "Large woody debris",
      overhanging_veg   = "Overhanging vegetation*",
      aquatic_veg       = "Aquatic vegetation*",
      cobble_substrate  = "Cobble substrate",
      boulder_substrate = "Boulder substrate",
      undercut_bank     = "Undercut bank",
      depth             = "Depth*",
      velocity          = "Velocity*",
      redd_total        = "Total Number of Redd*",
      redd_presence     = "Redds Nearby*",
      surface_turbidity = "Surface Turbulence"
    )
  )

ggplot(effects_df, aes(x = log_or, y = reorder(term, log_or))) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_point(size = 2.7) +
  geom_errorbarh(aes(xmin = log_or_low, xmax = log_or_high), height = 0.2) +
  labs(
    x = "Log odds ratio",
    y = NULL,
    title = "Effect sizes for predictors of steelhead presence",
    caption = "* represents model significance p<0.05"
  ) +
  theme_minimal(base_size = 13)

```

### Discussion

Steelhead presence in the Feather River was strongly influenced by local
habitat structure, spawning context, and broader spatial and seasonal
variation. The mixed‐effects logistic regression model that included
random intercepts for transect location and months March - May performed
very well, underscoring the importance of accounting for both site‐level
heterogeneity and seasonal dynamics when modeling steelhead occurrence.

Structural habitat features were the most consistent predictors of
steelhead presence. Overhanging vegetation, aquatic vegetation, and
small woody debris were all positively associated with occurrence,
indicating a strong preference for habitats that provide cover, refuge,
and hydraulic complexity. These findings align with the ecology of
juvenile steelhead, which typically select habitats that reduce
predation risk and energetic costs. In contrast, substrate variables and
large woody debris did not show strong independent effects once other
habitat features were considered.

Spawning context also played a key role. The presence of nearby redds
and higher total redd counts were both associated with increased
probabilities of steelhead presence, suggesting spatial coupling between
spawning and rearing habitats or shared reach‐scale conditions that
support both life stages. While these effects were strong, they were
estimated with greater uncertainty than structural habitat predictors.

Hydraulic variables had secondary influence. Velocity showed a negative
association with steelhead presence, indicating reduced use of
higher‐velocity habitats when suitable cover was lacking. Depth,
however, exhibited little effect, suggesting that depth alone is less
important than habitat complexity.

Substantial unexplained variation remained at both the site and seasonal
scales. Random‐effect estimates indicated that some transects
consistently supported higher baseline probabilities of steelhead
presence, while others were persistently low, reflecting the influence
of unmeasured reach‐scale factors such as connectivity, geomorphology,
or thermal conditions. Strong month‐level effects revealed pronounced
seasonal structure, with higher occurrence in spring and reduced
presence later in the season, consistent with outmigration patterns.
Together, these results highlight the combined importance of fine‐scale
habitat structure and broader spatial and temporal context in shaping
steelhead distribution.

### Supplementary Figures

**Conallin et al. 2014 - Figure 5**

```{r}
cover_vars <- c(
  "small_woody",
  "large_woody",
  "overhanging_veg",
  "undercut_bank",
  "aquatic_veg",
  "boulder_substrate",
  "cobble_substrate",
  "surface_turbidity"
)

site_var <- "channel_location"  

df_long <- log_reg_data |>
  mutate(presence = as.integer(count > 0))  |> 
  select(all_of(site_var), presence, all_of(cover_vars)) |> 
  mutate(across(all_of(cover_vars), ~ as.integer(.x > 0))) |> 
  pivot_longer(
    cols = all_of(cover_vars),
    names_to = "feature",
    values_to = "feature_present"
  )

feature_summary <- df_long |> 
  group_by(.data[[site_var]], feature) |> 
  summarise(
    n_all = n(),
    n_present_feature = sum(feature_present == 1, na.rm = TRUE),
    n_fish_present = sum(presence == 1, na.rm = TRUE),
    n_fish_present_with_feature = sum(presence == 1 & feature_present == 1, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  group_by(.data[[site_var]]) |> 
  mutate(
    HA = n_present_feature / n_all,
    HU = ifelse(n_fish_present > 0, n_fish_present_with_feature / n_fish_present, NA_real_),
    P  = HU / HA,
    HSI_raw = ifelse(is.finite(P), P / max(P, na.rm = TRUE), NA_real_)
  ) |> 
  ungroup()

plot_df <- feature_summary |> 
  select(.data[[site_var]], feature, HA, HU, HSI_raw) |> 
  pivot_longer(cols = c(HA, HU, HSI_raw), names_to = "panel", values_to = "value") |> 
  mutate(
    panel = recode(panel,
                   HA = "HA (availability)",
                   HU = "HU (utilization)",
                   HSI_raw = "HSI (preference)"),
    feature = str_replace_all(feature, "_", " ")
  )

ggplot(plot_df, aes(x = feature, y = value)) +
  geom_col(fill = "black", alpha = 0.8) +
  facet_grid(panel ~ .data[[site_var]], scales = "free_y") +
  labs(
    x = NULL,
    y = "Value (0–1)",
    title = "Steelhead - Habitat feature availability, use, and preference by channel location"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    strip.text = element_text(size = 9),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.line = element_blank(),
    plot.title = element_text(size = 12)
  )

ggsave(
  filename = here::here("mini-snorkel-analysis", "figures", "steelhead_hsi_lfc_hfc.png"),
  width = 12,
  height = 9,
  dpi = 300,
  bg = "white"
)
```


```{r}
cover_vars <- c(
  "small_woody",
  "large_woody",
  "overhanging_veg",
  "undercut_bank",
  "aquatic_veg",
  "boulder_substrate",
  "cobble_substrate",
  "surface_turbidity"
)

site_var <- "location"  

df_long <- log_reg_data |>
  mutate(presence = as.integer(count > 0))  |> 
  select(all_of(site_var), presence, all_of(cover_vars)) |> 
  mutate(across(all_of(cover_vars), ~ as.integer(.x > 0))) |> 
  pivot_longer(
    cols = all_of(cover_vars),
    names_to = "feature",
    values_to = "feature_present"
  )

feature_summary <- df_long |> 
  group_by(.data[[site_var]], feature) |> 
  summarise(
    n_all = n(),
    n_present_feature = sum(feature_present == 1, na.rm = TRUE),
    n_fish_present = sum(presence == 1, na.rm = TRUE),
    n_fish_present_with_feature = sum(presence == 1 & feature_present == 1, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  group_by(.data[[site_var]]) |> 
  mutate(
    HA = n_present_feature / n_all,
    HU = ifelse(n_fish_present > 0, n_fish_present_with_feature / n_fish_present, NA_real_),
    P  = HU / HA,
    HSI_raw = ifelse(is.finite(P), P / max(P, na.rm = TRUE), NA_real_)
  ) |> 
  ungroup()

plot_df <- feature_summary |> 
  select(.data[[site_var]], feature, HA, HU, HSI_raw) |> 
  pivot_longer(cols = c(HA, HU, HSI_raw), names_to = "panel", values_to = "value") |> 
  mutate(
    panel = recode(panel,
                   HA = "HA (availability)",
                   HU = "HU (utilization)",
                   HSI_raw = "HSI (preference)"),
    feature = str_replace_all(feature, "_", " ")
  )

ggplot(plot_df, aes(x = value, y = feature, color = feature)) +
  geom_jitter(height = 0.2, alpha = 0.5, size = 1.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  facet_wrap(~panel, ncol = 1) +
  scale_color_manual(
    values = c(
      "aquatic veg"        = "black",
      "boulder substrate"  = "black",
      "cobble substrate"   = "black",
      "fine substrate"     = "black",
      "gravel substrate"   = "black",
      "large wood"         = "black",
      "pool"               = "black",
      "riffle"             = "black",
      "run"                = "black"
    ),
    guide = "none"
  ) +
  labs(
    x = "Value (0–1)",
    y = NULL,
    title = "Steelhead habitat: habitat variability across locations",
    subtitle = "Points = locations; diamond = mean"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.background = element_rect(fill = "white", color = "black", linewidth = 0.5),
    strip.text = element_text(size = 10),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.line = element_blank(),
    plot.title = element_text(size = 12),
    plot.subtitle = element_text(size = 10, color = "grey40")
  )

ggsave(
  filename = here::here("mini-snorkel-analysis", "figures", "steelhead_hsi_locations.png"),
  width = 12,
  height = 9,
  dpi = 300,
  bg = "white"
)

```

```{r}
knitr::knit_exit()
```
