---
title: "Feather River - occupancy model"
author: "Ashley Vizek (FlowWest)"
date: "`r Sys.Date()`"
output: 
  html_document
---

# Overview

The goal of this document is to explore the use of an occupancy model for understanding important habitat charateristics for salmonids.

Feather River will be used as a case study location with the goal of creating a more streamlined workflow that could be applied elsewhere. There are two relevant datasets (1) mini snorkel data, and (2) the intermediate level ongoing snorkel survey. These data are utilizing different methodologies and part of these analyses will explore the differences between these two data collection methods and resulting habitat.

**This markdown focuses on occupancy modeling.**

```{r, include = F, message = F, warning = F}
library(tidyverse)
library(googleCloudStorageR)
knitr::opts_chunk$set(warning = F, message = F)

ongoing_location_lookup_raw <- read_csv(here::here("data", "feather_location_lookup.csv"))
ongoing_metadata_raw <- read_csv(here::here("data", "feather_snorkel_metadata.csv"))
ongoing_obs_raw <- read_csv(here::here("data", "feather_snorkel_observations.csv"))

# mini snorkel data
# TODO is there a table that associates the microhabitat plot with a lat/long?
# TODO the survey locations are missing coordinates
mini_snorkel_raw <- read_csv(here::here("data", "microhabitat_with_fish_observations.csv"))
location_raw <- read_csv(here::here("data", "survey_locations.csv"))
```

## Occupancy modeling

The goal here is to model presence/absence in a hierarchical Bayesian framework where there are repeat observations for a given site which incorporates sampling bias.

*Considerations in progress*

We need to consider if the mini-snorkel data is set up appropriately to use with this modeling framework. We would need to treat the reach/location as a site and define a time period where it is reasonable that fish would not have outmigrated. (There may be other ways to handle the change in observations overtime related to outmigration).
- The replicates were set up to be at the 1m2 level (36 replicates per site)
- There are 24 sites which may or may not be sampled multiple times

In this framing cover and turbidity would be affecting detection probability whereas site level variables like depth and velocity would be affecting occupancy. We could include cover at the site level as well. 

Reservations

- Average site level covariates - are these characterizing the site appropriately? is there another way to characterize the site and occupancy that would be a better fit?
- Is this the appropriate structuring for the data? We are treating each of the microhabitat sites as replicates.

Refer to: 

- Pascale's paper: https://onlinelibrary.wiley.com/doi/pdfdirect/10.1111/jfb.14143
- Mahardja 2021: https://escholarship.org/uc/item/9sp7r7q4 
- Another example: https://pubs.usgs.gov/of/2020/1100/ofr20201100.pdf

### Determine time period

- Need to determine the time period where the assumption that fish have not migrated out of the study area is likely.
- Salmon count drops by more than half in April so we should limit to observations in March.
- We could include all observations but also include a variable for month that affects occupancy.

```{r, include = F}
mini_snorkel_filtered <- mini_snorkel_raw |> 
  # associate location table so we can group by the reach if needed.
  # note that the location_revised field is still undergoing review with Ryon
  left_join(location_raw |> 
              select(location_table_id, location_revised)) |> 
  # Ryon thought 140 was better than 100. TODO look into this more. goal to filter to YOY
  filter(fl_mm < 140 | is.na(fl_mm)) |> 
  mutate(month = month(date))

mini_snorkel_filtered |> 
  group_by(month = month(date)) |> 
  summarize(count = sum(count, na.rm = T)) |> 
  ggplot(aes(x = month, y = count)) +
  geom_col()

mini_snorkel_filtered |> 
  mutate(month = month(date)) |> 
  select(location_revised, month) |> 
  distinct() |> 
  group_by(location_revised) |> 
  tally() |> 
  ggplot(aes(x = location_revised, y = n)) +
           geom_col()

mini_snorkel_filtered_march <- mini_snorkel_filtered |> 
  filter(month(date) == 3) 

# summarize the number of sites sampled - looks like there is one steep riffle for 3/13 and one for 3/15. this may be due to incorrect location cateogorization. for now lets call all location_revised as one single site

# summarize the number of replicates per site - as expected 35-36 for each

# decide on how to categorize cover
# woody debris, submerged aquatic, undercut, overhead - lets try this for now (select dominant)

mini_snorkel_filtered_march |> 
  distinct(location_revised, date) |> 
  arrange(location_revised)

mini_snorkel_filtered_march |> 
  distinct(location_revised, transect_code) |> 
  group_by(location_revised) |> 
  tally()
```

```{r, include = F}

# processing cover by: 1. joining large/small woody  2. joining the two overhead levels 3. assigning highest percent 4. assigned 0 if NA

mini_snorkel_filtered_cover <- mini_snorkel_filtered_march |> 
  select(location_revised, transect_code, percent_small_woody_cover_inchannel, percent_large_woody_cover_inchannel, percent_submerged_aquatic_veg_inchannel, percent_undercut_bank, percent_cover_half_meter_overhead, percent_cover_more_than_half_meter_overhead) |> 
  distinct() |> 
  mutate(woody = percent_small_woody_cover_inchannel + percent_large_woody_cover_inchannel,
         overhead = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |> 
  select(-c(percent_small_woody_cover_inchannel, percent_large_woody_cover_inchannel, percent_cover_half_meter_overhead, percent_cover_more_than_half_meter_overhead)) |> 
  rename(aquatic = percent_submerged_aquatic_veg_inchannel,
         undercut = percent_undercut_bank) |> 
  pivot_longer(cols = c(woody, aquatic, undercut, overhead)) |> 
  group_by(location_revised, transect_code) |> 
  slice_max(value, with_ties = F) |> 
  mutate(cover_type = ifelse(value == 0, "no cover", name)) |> 
  select(-c(name, value))

full_mini_snorkel_data <- mini_snorkel_filtered |> 
  left_join(mini_snorkel_filtered_cover)

# This can be used to clean the data if needed. If there are ties they happen when there are 0s
# so for now adjusting to remove ties from slice_max. Remove this when ready to clean
mini_snorkel_filtered_march_cover |> 
  group_by(micro_hab_data_tbl_id, location_table_id, transect_code, date) |> 
  summarize(sum = sum(value),
            n = length(transect_code)) |> 
  filter(n > 1)

```

```{r, include = F}
library(unmarked)
# sets up mini snorkel data for occupancy model
# habitat variables of interest based on literature
# depth, velocity, substrate, cover

# site level habitat variables
# we would need to take an average for the entire site based on micro level information
# TODO are there better site level variables to use?
# channel geomorphic is collected at microhab level it appears
# use average depth and velocity for now
# including cover by taking average percent of types and then applying same dominant methodology. could do this in a different way

average_site_cover <- mini_snorkel_filtered |> 
  select(location_revised, transect_code, percent_small_woody_cover_inchannel, percent_large_woody_cover_inchannel, percent_submerged_aquatic_veg_inchannel, percent_undercut_bank, percent_cover_half_meter_overhead, percent_cover_more_than_half_meter_overhead) |> 
  distinct() |> 
  group_by(location_revised) |> 
  summarise(percent_small_woody_cover_inchannel = mean(percent_small_woody_cover_inchannel),
            percent_large_woody_cover_inchannel = mean(percent_large_woody_cover_inchannel),
            percent_cover_half_meter_overhead = mean(percent_cover_half_meter_overhead),
            percent_cover_more_than_half_meter_overhead = mean(percent_cover_more_than_half_meter_overhead),
            percent_submerged_aquatic_veg_inchannel = mean(percent_submerged_aquatic_veg_inchannel),
            percent_undercut_bank = mean(percent_undercut_bank)) |> 
  mutate(woody = percent_small_woody_cover_inchannel + percent_large_woody_cover_inchannel,
         overhead = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |> 
  select(-c(percent_small_woody_cover_inchannel, percent_large_woody_cover_inchannel, percent_cover_half_meter_overhead, percent_cover_more_than_half_meter_overhead)) |> 
  rename(aquatic = percent_submerged_aquatic_veg_inchannel,
         undercut = percent_undercut_bank) |> 
  pivot_longer(cols = c(woody, aquatic, undercut, overhead)) |> 
  group_by(location_revised) |> 
  slice_max(value, with_ties = F) |> 
  rename(percent_cover = value) |> 
  select(-c(name))
 
average_site_level_covariates <- full_mini_snorkel_data |> 
  select(-c(fish_data_id, species, date, channel_geomorphic_unit, fl_mm, count, dist_to_bottom, focal_velocity)) |> 
  distinct() |> # make sure we just have one row for each quadrat (if multiple types of fish/fl observed would be repeat rows)
    mutate(
         mean_velocity = mean(velocity), # scale variables
         mean_depth = mean(depth),
         sd_velocity = sd(velocity),
         sd_depth = sd(depth),
         velocity = ((velocity - mean_velocity)/sd_velocity),
         depth = ((depth - mean_depth)/sd_depth)) |> 
  pivot_longer(cols = c(depth:velocity), names_to = "parameter", values_to = "value") |> 
  group_by(location_revised, parameter) |> 
  summarize(value = mean(value, na.rm = T)) |>  # decided to do mean because there are a lot of zeros; we could also choose min or max
  pivot_wider(id_cols = location_revised, values_from = "value", names_from = "parameter") |> 
  left_join(average_site_cover)

# quadrat level fish observations variables
quadrat_fish_obs <- full_mini_snorkel_data |> 
  # decided not to do species filtering for now
  # filter(species == "Chinook salmon" | grepl("Steelhead", species) | is.na(species)) |> 
  select(location_revised, transect_code, count) |> 
  group_by(location_revised, transect_code) |> 
  summarize(count = sum(count, na.rm = T)) |> 
  mutate(count = ifelse((is.na(count) | count == 0), 0, 1)) |> 
  group_by(location_revised) |> 
  mutate(label = row_number()) 

quadrat_fish_obs_wide <- quadrat_fish_obs |> 
  pivot_wider(id_cols = location_revised, names_from = "label", values_from = "count")

# preparing covariates for observations. focal velocity and depth will vary by species obs. we can take the average or we can use the depth and velocity associated with the quadrat
# decided just to use cover for now
quadrat_cover <- quadrat_fish_obs |> 
              select(-count) |> 
  left_join(mini_snorkel_filtered_march_cover) |> 
  mutate(cover_type = factor(cover_type))

cover <- quadrat_cover |> 
  pivot_wider(id_cols = location_revised, names_from = "label", values_from = "cover_type")

quadrat_turbidity <- quadrat_fish_obs |> 
  select(-count) |> 
  left_join(full_mini_snorkel_data |> 
              group_by(location_revised, transect_code) |> 
              summarize(turbidity = mean(surface_turbidity, na.rm = T))) |> 
  mutate(mean_turbidity = mean(turbidity),
         sd_turbidity = sd(turbidity),
         turbidity = ifelse(mean_turbidity==0 & sd_turbidity==0, 0, ((turbidity - mean_turbidity)/sd_turbidity))) |> 
  select(-c(mean_turbidity, sd_turbidity))

turbidity <- quadrat_turbidity |> 
  pivot_wider(id_cols = location_revised, names_from = "label", values_from = "turbidity")


# decided to only use cover for now
# depth <- quadrat_hab_vars  |> 
#   select(location_table_id, label, depth) |> 
#   ungroup() |> 
#   mutate(label = paste0("depth.", label),
#          mean= mean(depth),
#          sd = sd(depth),
#          depth = ((depth - mean)/sd)) |> 
#   pivot_wider(id_cols = location_table_id, names_from = "label", values_from = "depth")

# velocity
# velocity <- quadrat_hab_vars  |> 
#   select(location_table_id, label, velocity) |> 
#   ungroup() |> 
#   mutate(label = paste0("velocity.", label),
#          mean = mean(velocity, na.rm = T),
#          sd = sd(velocity, na.rm = T),
#          velocity = ((velocity - mean)/sd)) |> 
#   pivot_wider(id_cols = location_table_id, names_from = "label", values_from = "velocity")

# # large sub
# large_sub <- quadrat_hab_vars  |> 
#   select(location_table_id, label, percent_large_substrate) |> 
#   ungroup() |> 
#   mutate(label = paste0("substrate.", label),
#          mean = mean(percent_large_substrate),
#          sd = sd(percent_large_substrate),
#          percent_large_substrate = ((percent_large_substrate - mean)/sd)) |> 
#   pivot_wider(id_cols = location_table_id, names_from = "label", values_from = "percent_large_substrate")

# instream
# instream <- quadrat_hab_vars  |> 
#   select(location_table_id, label, percent_instream_cover) |> 
#   ungroup() |> 
#   mutate(label = paste0("instreamcover.", label),
#          mean = mean(percent_instream_cover),
#          sd = sd(percent_instream_cover),
#          percent_instream_cover = ((percent_instream_cover - mean)/sd)) |> 
#   pivot_wider(id_cols = location_table_id, names_from = "label", values_from = "percent_instream_cover")

# overhead
# overhead <- quadrat_hab_vars  |> 
#   select(location_table_id, label, percent_overhead_cover) |> 
#   ungroup() |> 
#   mutate(label = paste0("overheadcover.", label),
#          mean = mean(percent_overhead_cover),
#          sd = sd(percent_overhead_cover),
#          percent_overhead_cover = ((percent_overhead_cover - mean)/sd)) |> 
#   pivot_wider(id_cols = location_table_id, names_from = "label", values_from = "percent_overhead_cover")

y <- quadrat_fish_obs_wide[,2:37]
siteCovs <- average_site_level_covariates[, c("depth", "velocity", "percent_cover")]
obsCovs <- list(cover = cover[,2:37],
                turbidity = turbidity[,2:37])

dat <- unmarkedFrameOccu(y = y, siteCovs = siteCovs, obsCovs = obsCovs)
summary(dat)
plot(dat)
# example datasets within unmarked package
# https://cran.r-project.org/web/packages/unmarked/vignettes/unmarked.html

# wt <- read.csv(system.file("csv","widewt.csv", package="unmarked"))
# y <- wt[,2:4]
# siteCovs <-  wt[,c("elev", "forest", "length")]
# obsCovs <- list(date=wt[,c("date.1", "date.2", "date.3")],
#     ivel=wt[,c("ivel.1",  "ivel.2", "ivel.3")])
# wt <- unmarkedFrameOccu(y = y, siteCovs = siteCovs, obsCovs = obsCovs)
# summary(wt)
# wt <- read_csv(system.file("csv", "frog2001pcru.csv", package = "unmarked"))
# pcru <- csvToUMF(system.file("csv","frog2001pcru.csv", package="unmarked"),
#                  long = TRUE, type = "unmarkedFrameOccu")
```

```{r, eval = F, include = F}
# TODO need to better understand detection v occupancy
# Considerations - should this be an open model because fish could be migrating downstream

# ~ detection ~ occupancy

# Null model
fm1 <- occu(~1 ~1, dat)
fm1
# Occupancy probability is 0.99 (nearly 1) and detection probability is 0.17
backTransform(fm1, type = "state")
backTransform(fm1, type = "det")

# Detection is modeled with depth and velocity
fm2 <- occu(~depth+velocity ~1, dat)
fm2


# The expected probability that a site was occupied is 0.99. This estimate applies to the hypothetical population of all possible sites, not the sites found in our sample. For a good discussion of population-level vs finite-sample inference, see Royle and Dorazio (2008) page 117. Note also that finite-sample quantities can be computed in unmarked using empirical Bayes methods as demonstrated at the end of this document.
backTransform(fm2, type = "state")

# Thus, we can say that the expected probability of detection was 0.17 when depth and velocity are fixed at their mean value. 
backTransform(linearComb(fm2, coefficients = c(1,0,0), type = 'det'))
# predictions increase as velocity increases but never above 0.3
newData <- data.frame(depth = 0, velocity = -2:2)
round(predict(fm2, type = 'det', newdata = newData, appendData=TRUE), 2)
# predictions decrease as depth decreases but never above 0.28
newData <- data.frame(depth = -2:2, velocity = 0)
round(predict(fm2, type = 'det', newdata = newData, appendData=TRUE), 2)
# newData <- data.frame(depth = -2:2, velocity = 0, instream_cover = 0)
# newData <- data.frame(depth = 0, velocity = 0, instream_cover = -2:2)

# Detection is modeled with cover and turbidty
# highest prob of detection is with overhead and low turbidity
# high turbidity decreases prob of detection
fm3 <- occu(~cover+turbidity ~1, dat)
fm3 # overhead is the only significant variable
newData <- data.frame(cover = c("aquatic","no cover", "overhead", "woody", "undercut"),
                      turbidity = 0)
newData <- data.frame(cover = c("aquatic","no cover", "overhead", "woody", "undercut"),
                      turbidity = -2)
newData <- data.frame(cover = c("aquatic","no cover", "overhead", "woody", "undercut"),
                      turbidity = 2)
predict(fm3, type = 'det', newdata = newData, appendData=TRUE)

# Detection is modeled with turbidity
# If turbidity held at mean detection probability is 0.17, if decreased then 0.20, if increased then 0.14

fm_turbid <- occu(~turbidity ~1, dat)
fm_turbid
backTransform(linearComb(fm_turbid, coefficients = c(1,-2), type = 'det')) # this line of code allows us to see how detection probability changes when adjust turbidity

fm_cover <- occu(~cover ~1, dat)
fm_cover
newData <- data.frame(cover = c("aquatic","no cover", "overhead", "woody", "undercut"))
predict(fm_cover, type = 'det', newdata = newData, appendData=TRUE)

# Detection is modeled with cover and occupancy is modeled with depth and velocity
fm4 <- occu(~cover ~depth+velocity, dat)
fm4
# how do predictions change for different cover types
newData <- data.frame(cover = c("aquatic","no cover", "overhead", "woody", "undercut"))
predict(fm4, type = 'det', newdata = newData, appendData=TRUE)
# how do prediction of occupancy change based on depth/velocity
newData <- data.frame(depth = -2:2, velocity = 0)
newData <- data.frame(depth = 0, velocity = -2:2)
predict(fm4, type = 'state', newdata = newData, appendData=TRUE)

# Detection modeled with cover and turbidity, occupancy modeled by depth, velocity and cover
full_mod <- occu(~cover+turbidity ~depth+velocity+percent_cover, dat)
full_mod
```
