---
title: "feather river redd data qc -  continuation"
date: "2024-11-06"
output: 
  html_document:
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

library(tidyverse)
library(lubridate)
library(googleCloudStorageR)
library(ggplot2)
library(scales)
library(dplyr)
library(leaflet)
library(RColorBrewer)
library(viridis)
library(readxl)
library(janitor)
library(sf)
library(openxlsx)
library(readxl)
library(leaflet)
library(readr)
```

This markdown is a continuation of mapping_qc.Rmd. Based on the issues found on that markdown, the issues will be fixed here.
The goal of this markdown is to clean up the location names from Feather River redd survey data.

Summary of work that has been done so far:

  * *redd_initial_qc* - data pull and initial qc: 
    * raw data is pulled in and combined
    * variable names are cleaned
    * negative velocity values are set to NA
    * pot_depth_m >= 40 is set to NA
    * "type" is removed
    * 2023 number_redds set all to 1 
    * note that 7 data entries in the raw data that had "0" in number_redds, those were set to 1
    * Data file output: "data-raw/combined_raw_data.csv"
    
  * *mapping-qc-exploration*: exploration script - looks into latitude longitude and identifies issues. 
    
  * *redd_qc_cont*: fixed latitude/longitude provided by Chris Cook, these changes address the issues found on mapping-qc-exploration.Rmd.

```{r, include=FALSE}
# redd survey points
redd_data <- read.csv("data-raw/combined_raw_data.csv")

# surveying sites from gis data map
gdb_path <- "data-raw/shapefiles/b105_redd_survey_locations_maps_2021.gdb"

layers <- st_layers(gdb_path)$name
```


```{r, include=FALSE}
# there are a few data entries that have 0 for either latitude or longitude
# filtering out those lat/long that can't be plotted
redd_data |> 
  filter(latitude == 0 |
         longitude == 0) |>  #checking that they are 47 rows that will be fixed
  glimpse()

data_zero_removed <- redd_data |> 
    filter(latitude != 0, #filtering out those 47 rows
         longitude != 0) |>
  glimpse()
  
# •	2022 – 1 data entry
# •	2020 – 4 data entries 
# •	2019 – 10 data entries
# •	2018 – 6 data entries
# •	2015 – 18 data entries
# •	2014 – 8 data entriesca


# Chris shared the correct lat/long for some of these records, and decided to delete some of the entries
fixed_zeros <- read.csv("data-raw/qc-processing-files/Faulty_lat_long_records_edits_Chris_111224.csv") |> # csv with Chris' notes
  select(-1) |> 
  mutate(Northing = str_trim(Northing)) |>
  filter(Northing != "Erase this data") |> #erasing the 6 data entries he indicates to delete
  select(-c(latitude, longitude, Northing, Easting, X.1)) |>
  clean_names() |>
  glimpse()

#join fixed zero values, and deleted data entries 
redd_data_fixed_zeros <- data_zero_removed |> 
  bind_rows(fixed_zeros) |> 
  glimpse()

```

```{r, include=FALSE}
# filtering the points that off channel
off_channel_removed <- redd_data_fixed_zeros |>
  filter(
!(
  (location == "lower auditorium" & date == "2014-09-18" & survey_wk == "2-2" & file_number == "66") |
(location == "lower auditorium" & date == "2015-10-29" & survey_wk == "7-3" & file_number == "5") |
(location == "lower auditorium" & date == "2015-10-22" & survey_wk == "6-4" & file_number == "44") |
(location == "top of auditorium" & date == "2015-11-24" & survey_wk == "11-1" & file_number == "44") |
(location == "upper hatchery riffle" & date == "2014-10-22" & survey_wk == "7-2" & file_number == "37") |
(location == "table mountain" & date == "2015-09-22" & survey_wk == "2-2" & file_number == "10") |
(location == "top of auditorium" & date == "2014-10-20" & survey_wk == "7-1" & file_number == "88") |
(location == "top of auditorium" & date == "2014-10-14" & survey_wk == "6-1" & file_number == "84") |
(location == "lower auditorium" & date == "2015-10-19" & survey_wk == "6-1" & file_number == "16") |
(location == "middle auditorium" & date == "2015-10-22" & survey_wk == "6-4" & file_number == "106") |
(location == "upper hatchery" & date == "2015-10-28" & survey_wk == "7-2" & file_number == "17") |
(location == "mid auditorium" & date == "2014-11-13" & survey_wk == "8-2" & file_number == "174") |
(location == "moe's side channel" & date == "2015-10-09" & survey_wk == "4-4" & file_number == "87") |
(location == "moe's side channel" & date == "2015-10-30" & survey_wk == "7-4" & file_number == "87") |
(location == "lower auditorium" & date == "2014-10-01" & survey_wk == "4-2" & file_number == "64") |
(location == "upper hatchery riffle" & date == "2014-10-02" & survey_wk == "4-3" & file_number == "31")
)
)|>
glimpse()

#reding in Chris' file with corrections
fixed_off_channel <- read.csv("data-raw/qc-processing-files/Off_channel_records_edit_Chris_110524.csv") |> 
  glimpse()

fixed_off_channel <- fixed_off_channel |> 
  clean_names() |> 
  select(-c(latitude, longitude, corrected_utm_easting, corrected_utm_northing, x)) |>
  rename(latitude = "corrected_latitude",
         longitude = "corrected_longitude") |>
  glimpse()

redd_locations_fixed <- off_channel_removed |> 
  bind_rows(fixed_off_channel) |> 
  glimpse()
```


```{r}
leaflet(redd_locations_fixed) |> 
  addTiles() |> 
  addCircleMarkers(
    ~longitude, ~latitude,
    popup = ~paste0("Number of Redds: ", number_redds, "<br>Lat: ", latitude,"<br>Long: ", longitude),
    radius = 2,  # Adjust radius as needed
    color = "red",
    # label = ~as.character(location),
    # labelOptions = labelOptions(noHide = TRUE, textOnly = TRUE)
  )
```

### Looking into unique location names

  * There are 81 unique names
  
```{r}
unique(redd_locations_fixed$location)
```

```{r, include=FALSE}
# removing spaces
redd_locations_fixed$location <- gsub(" ", "", redd_locations_fixed$location)

# Display the modified vector
print(unique(redd_locations_fixed$location))

```


Survey location names - 
The following names below are survey locations names provided by Feather River team. These location names are what they have on their reports and how they categorize the locations where surveys are conducted. Redd data locations will be matched to these names

```{r, echo=FALSE, message=FALSE}
# Names from survey location provided by Chris Cook
# Great Western, Palm, Top of Hatchery, G95 Main and Lower Vance East are not included on locations provided but we asked Chris and we should keep them
reference_locations <- c(
  "UpperHatchery", "UpperCottonwood", "Hatchery", "Cottonwood",
  "TopofAuditorium", "MoesSideChannel", "UpperAuditorium", "MiddleAuditorium",
  "LowerAuditorium", "HatcheryPipe", "LowerTableMountain", "TableMountain",
  "BelowLowerAuditorium", "Bedrock", "TrailerPark", "UpperMathews",
  "Mathews", "UpperRobinson", "LowerRobinson", "Steep",
  "SteepSideChannel", "Weir", "Eye", "EyeSideChannel",
  "Gateway", "Aleck", "HatcherySideChannel", "VanceEast",
  "VanceWest", "BigHoleEast", "BigHoleWest", "BelowBigHoleEast",
  "G95WestSideChannel", "G95EastSideChannelTop", "G95EastSideChannelBottom", "UpperHour",
  "LowerHour", "Keister", "Goose", "Big",
  "LowerBig", "BigBar", "UpperMcFarland", "LowerMcFarland",
  "Developing", "GreatWestern", "Palm", "TopofHatchery", "G95main", "LowerVanceEast"
)

reference_locations <- tolower(reference_locations)
table_data <- data.frame(
  ID = seq_along(reference_locations), 
  Name = reference_locations
)
table_data <- table_data[order(table_data$Name), ]

# Generate the sorted table
knitr::kable(table_data, col.names = c("ID", "Location Name"))
```

Identifying redd locations that do not match survey location names

```{r, echo=FALSE}
invalid_locations <- redd_locations_fixed$location[!redd_locations_fixed$location %in% reference_locations]

# Display invalid locations
unique_invalid_locations <- unique(invalid_locations)
print(unique_invalid_locations)
```

```{r, include=FALSE}
# renaming those that are just miss spelled 
redd_observations_clean_1 <- redd_locations_fixed |> 
  mutate(location = case_when(
    location == "aleckriffle" ~ "aleck",
    location == "bedrockriffle" ~ "bedrock",
    location == "bigriffle" | location ==	"bigbarriffle" ~	"bigbar",
    location == "g-95eastsidechannelbottom"	| location == "g95eastbottom"	~ "g95eastsidechannelbottom",
    location == "g-95easttop" | location == "g95easttop"	~ "g95eastsidechanneltop",
    location == "g-95westsidechannelbottom" | location ==	"g95west" | location ==	"g95westbottom" ~	"g95westsidechannel",
    location == "gooseriffle"	~	"goose",
    location == "hatcheryriffle" ~ "hatchery", # unsure relationship
    location == "hatcherysidechannelupper" ~	"hatcherysidechannel",
    location == "lowerbigriffle" ~	"lowerbig",
    location == "matthews" ~	"mathews",
    location == "midauditorium"| location == "midhour" ~	"middleauditorium", # unsure about midhour relationship
    location == "moe'ssidechannel" ~ "moessidechannel",
    location == "steepriffle"	~	"steep", # unsure
    location == "steepside" ~ "steepsidechannel", #unsure
    location == "topofaudtorium" ~ "topofauditorium",
    location == "upperhatcheryriffle" ~	"upperhatchery",
    location == "uppermatthews" ~	"uppermathews",
    location == "weirriffle"	~	"weir",
    location == "developingriffle" ~ "developing",
    location == "gatewaysidechannel" | location == "gatewaymainchannel" ~ "gateway",
    location == "upperbigholeeast" ~ "bigholeeast",
    location == "g-95main" ~ "g95main",
    location == "palmriffle" ~ "palm",
    TRUE ~ location)) 

print(unique(redd_observations_clean_1$location))
```

Redd location names that do not match survey locations, even after naming cleaning

```{r, echo=FALSE}
# checking for the locations that still dont match reference names, after cleaning above
print(unique(redd_observations_clean_1$location[!redd_observations_clean_1$location %in% reference_locations]))
```


```{r, include=FALSE}
names_for_check <- redd_observations_clean_1 |> 
   filter(!location %in% reference_locations)  # Keep only the rows with inconsistent location names
# not_consistent_names <- st_as_sf(not_consistent_names, coords = c("longitude", "latitude"), crs = 4326)
```

Reference map to check where redds are located geographically

```{r echo=FALSE, warning=FALSE}
# pulling layers from survey location map for reference
gdb_path <- "data-raw/shapefiles/b105_redd_survey_locations_maps_2021.gdb"
layers <- st_layers(gdb_path)$name

name_check_map <- leaflet() |> 
  addTiles() |> 
  setView(lng = -121.5695, lat = 39.3665, zoom = 10)

for (layer_name in layers) {
  shapefile <- st_read(dsn = gdb_path, layer = layer_name)
  shapefile <- st_transform(shapefile, crs = 4326) |> 
    st_zm(drop = TRUE, what = "ZM")
  
name_check_map <- name_check_map |> 
    addPolygons(
      data = shapefile,
      fillColor = "lightblue",  
      color = "darkblue",
      weight = 2,               
      opacity = 1.0,            
      fillOpacity = 0.5,        
      group = layer_name,
      popup = paste("Layer Name: ", layer_name)
    )
}

locations <- unique(names_for_check$location)
name_check_map <- name_check_map |> 
  addCircleMarkers(data = names_for_check, 
                   ~longitude, ~latitude, 
                   label = ~paste(location),
                   popup = ~paste("Location:", location, "<br>Date:", date, "<br>Latitude:", latitude, "<br>Longitude:",
                                  longitude, "<br>Survey Week:", survey_wk),
                   group = ~location,
                   radius = 2,  
                   color = "red")

name_check_map <- name_check_map |> 
  addLayersControl(
    overlayGroups = c(locations), # Add chek boxes for the location names 
    options = layersControlOptions(collapsed = FALSE)
  )

# Display the map
name_check_map
```

### Notes / observations form map:

  * Great Western, Palm, Top of Hatchery, G95 Main and Lower Vance East are not included on location shapefiles provided, but we asked Chris and we should keep them
  * hour: most points are located near "LowerHour", however two are within "UpperHour"
  * palm / palmriffle are not within any survey location - it was not modified
  * g95main: There are two points that look like outliers from other points - we asked Chris and poins have been updated accordingly below
  * lowervanceeast: 3 points are within "BigHoleEast" (these were modified to BigHoleEast), all others are off any survey location - south of "VanceEast", north of "BigHoleEast"
  * gatewaymainchannel: located on survey location Gateway, modified
  * upperbigholeeast: modified to BigHoleEast
  * moe'sditch: modified 
  * upper mathews: 
  
Modifications - all points that fall within a survey location will be modified to match to that name

```{r, include=FALSE}
redd_observations_clean_2 <- redd_observations_clean_1 |>
  mutate(location = case_when(
     location == "hour" & date == "2015-11-19" & latitude == "39.4225537735696" & longitude == "-121.624800693264" ~ "upperhour",
     location == "hour" & date == "2015-11-19" & latitude == "39.4224821992187" & longitude == "-121.624878645819" ~ "upperhour",
     location == "hour" ~ "lowerhour",
     location == "upperhoureast" & date == "2014-10-09" & latitude == "39.4166671193809" & longitude == "-121.627212180498"
    ~ "lowerhour",
    location == "upperhoureast" & date == "2014-10-09" & latitude == "39.4159674094301" & longitude == "-121.627254551989" 
    ~ "lowerhour",
    location == "developingriffle" ~ "developing",
    location == "moe'sditch" & !(latitude == "39.5156950186919" & longitude == "-121.559350861003") ~ "moessidechannel",
    location == "moe'sditch" & latitude == "39.5156950186919" & longitude == "-121.559350861003" ~ "middleauditorium",
    location == "lowervanceeast" & date == "2017-11-14" & latitude == "39.439059494603" & longitude == "-121.635427259305" ~ "bigholeeast",
    location == "lowervanceeast" & date == "2017-11-14" & latitude == "39.4388562088935" & longitude == "-121.635448923638" ~ "bigholeeast",
    location == "lowervanceeast" & date == "2017-11-14" & latitude == "39.4388504794907" & longitude == "-121.635372994652" ~ "bigholeeast",
    location == "g95main" & date == "2015-11-19" & latitude == "39.4394792142928" & longitude == "-121.635254377528" ~ "bigholeeast",
    location == "g95main" & date == "2015-11-19" & latitude == "39.4371081082104" & longitude == "-121.636483707754" ~  "bigholeeast",
    location == "uppermathews" & date == "2023-10-25" & latitude == "39.5157891" & longitude == "-121.55643006" ~ "upperauditorium", # TODO from here below were modified on 01/08/25 - might want to confirm with Chris that modifications are ok. Names were changed based on their actual location
    location == "uppermathews" & date == "2023-10-25" & latitude == "39.51585525" & longitude == "-121.55658786" ~ "upperauditorium",
    location == "uppermathews" & date == "2023-10-25" & latitude == "39.51600506" & longitude == "-121.55655445" ~ "upperauditorium",
    location == "uppermathews" & date == "2023-10-25" & latitude == "39.51589657" & longitude == "-121.55733036" ~ "upperauditorium",
    location == "uppermathews" & date == "2023-10-19" & latitude == "39.51572598" & longitude == "-121.5576469" ~ "upperauditorium",
    location == "uppermathews" & date == "2023-10-19" & latitude == "39.51570684" & longitude == "-121.55787447" ~ "upperauditorium",
    location == "trailerpark" & date == "2023-09-26" & latitude == "39.51600036" & longitude == "-121.55916329" ~ "moessidechannel",
    location == "tablemountain" & date == "2022-11-22" & latitude == "39.496410384211" & longitude == "-121.579623191602" ~ "trailerpark",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4176998620939" & longitude == "-121.627223819039" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.417653889353" & longitude == "-121.627343452105" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4174661545398" & longitude == "-121.62741308081" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4174294735085" & longitude == "-121.627411511883" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4173586922004" & longitude == "-121.627554538371" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.41572855345" & longitude == "-121.626946544251" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4157329027447" & longitude == "-121.627060052528" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4156719809222" & longitude == "-121.627129956771" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4156289839513" & longitude == "-121.627133750753" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4156255231564" & longitude == "-121.627261165499" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4098841591031" & longitude == "-121.625766739867" ~ "lowerhour",
    location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4192277254991" & longitude == "-121.626608190798" ~ "lowerhour",
    TRUE ~ location
  ))


# pending points to ask Chris - 
# pending <- redd_observations_clean_2 |> 
#   filter(location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4192277254991" & longitude == "-121.626608190798" |
#          location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4098841591031" & longitude == "-121.625766739867") |> 
#   glimpse()


print(unique(redd_observations_clean_2$location))

print(unique(redd_observations_clean_2$location[!redd_observations_clean_2$location %in% reference_locations]))
# all redd locations now fall within reference locations!
```

```{r, include=FALSE}
#pending as of 01/08/25 --- RESOLVED
# redd_questions <- redd_observations_clean_1 |> 
#   filter(location == "uppermathews" & date == "2023-10-25" & latitude == "39.5157891" & longitude == "-121.55643006" |
#     location == "uppermathews" & date == "2023-10-25" & latitude == "39.51585525" & longitude == "-121.55658786" |
#     location == "uppermathews" & date == "2023-10-25" & latitude == "39.51600506" & longitude == "-121.55655445" |
#     location == "uppermathews" & date == "2023-10-25" & latitude == "39.51589657" & longitude == "-121.55733036" |
#     location == "uppermathews" & date == "2023-10-19" & latitude == "39.51572598" & longitude == "-121.5576469" | 
#     location == "uppermathews" & date == "2023-10-19" & latitude == "39.51570684" & longitude == "-121.55787447"| 
#     location == "trailerpark" & date == "2023-09-26" & latitude == "39.51600036" & longitude == "-121.55916329" | 
#     location == "tablemountain" & date == "2022-11-22" & latitude == "39.496410384211" & longitude == "-121.579623191602" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4176998620939" & longitude == "-121.627223819039" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.417653889353" & longitude == "-121.627343452105" |
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4174661545398" & longitude == "-39.4174661545398" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4174294735085" & longitude == "-121.627411511883" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4173586922004" & longitude == "-121.627554538371" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.41572855345" & longitude == "-121.626946544251" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4157329027447" & longitude == "-121.627060052528" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4156719809222" & longitude == "-121.627129956771"| 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4156289839513" & longitude == "-121.627133750753" | 
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4156255231564" & longitude == "-121.627261165499" |
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4174661545398" & longitude == "-121.627261165499" |
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4192277254991" & longitude == "-121.626608190798" |
#     location == "middleauditorium" & date == "2014-11-07" & latitude == "39.4098841591031" & longitude == "-121.625766739867") |>
#   glimpse()
# 
# write.csv(redd_questions, "data-raw/redd_questions_0125.csv")
```

Listing redd location names that remain unmatched to any survey locations (should be 0)

```{r, echo=FALSE}
# another check
names_for_check_2 <- redd_observations_clean_2 |> 
   filter(!location %in% reference_locations) 
  

unique(names_for_check_2$location)
```


```{r, include=FALSE}
redd_observations_wnames <- redd_observations_clean_2 |>
  mutate(date = case_when(
    str_detect(date, "/") ~ as.Date(date, format = "%m/%d/%Y"), # fix dates
    TRUE ~ as.Date(date))) |>
  select(-file_number) |>  # keeping survey week and location
  view()
```

```{r}
# removing fields that are not relevant to publish
redd_observations <- redd_observations_wnames |> 
  select(-c(survey_wk)) # confirm that we are keeping location
```


```{r, include=FALSE}
# save clean data - keeping survey week to use as reference (identify when 0 redds were observed)
write.csv(redd_observations_wnames, "data-raw/qc-processing-files/redd_observations_survey_wk.csv", row.names = FALSE)

# save clean data
write.csv(redd_observations, "data/redd_observations.csv", row.names = FALSE)
```


```{r, include=FALSE}
read.csv("data/redd_observations.csv") |> glimpse()
```


