---
title: "Feather River Habitat Suitability Index Development"
author: "Maddee Rubenson (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

### Adapted Mark Gard 1998 HSC process to Cover and Substrate

*adapted from <https://github.com/FlowWest/va-feather-river-analysis/blob/mini-snorkel/analysis/feather_river_mini_snorkel_hsi_analysis_exploration.md>*

[Mark Gard’s 1998 paper](https://personales.upv.es/fmcapel/data/Technique%20for%20Adjusting%20Spawning%20Depth%20Habitat%20Utilization%20Curves%20for%20Availability.pdf) identified a method to developing HSC for presence and absence of redds in different substrate classes. I will follow the same methodology using fish presence/absence and total fish count instead of redds and also apply to cover.

1.  Determine number of fish with each substrate-size class

2.  Calculate the proportion of fish with each substrate size class

3.  Calculate the HSC value for each substrate size class by dividing the proportion of fish in each substrate class by the proportion of fish with the most frequent substrate class

To convert from percent to discrete:

-   Cover decision: 20% or greater is considered present, which is consistent with the Strategic Plan (table 27, TODO include citation)

Cover includes (table 28 of the Strategic Plan):

-   small woody debris

-   large woody debris

-   boulder

-   cobble

-   grass/herbaceous - emergent vegetation

-   aquatic vegetation - non-emergent

-   willow and other riparian vegetation

-   undercut bank

-   overhanging vegetation

### Case Study - Feather River Mini Snorkel Data

Data was collected as percent cover for the following cover types. This crosswalk was applied to create consistent cover types across the mini snorkel data.

| Mini Snorkel Data Type                                                          | Applied Data Type                    |
|------------------------------------------------|------------------------|
| percent large woody debris                                                      | large woody cover                    |
| percent small wood                                                              | small woody cover                    |
| percent boulder substrate                                                       | boulder                              |
| percent cobble substrate                                                        | cobble                               |
| percent undercut bank                                                           | undercut bank                        |
| percent submerged aquatic vegetation inchannel                                  | aquatic vegetation                   |
| percent cover half meter overhead + percent cover more than half meter overhead | overhanging vegetation               |
| NA                                                                              | grass/herbaceous                     |
| NA                                                                              | willow and other riparian vegetation |

**Decisions**

-   The HSI's were developed for fish presence or absence (1/0) regardless of how many fish were present. This was decided to eliminate potential challenges in counting number of fish present and with the assumption that if a fish was present, the habitat is suitable for other fish to also be present.

**Discussion**

Substrate:

-   Sand has a stronger HSI when looking at number of fish vs. presence/absence

-   Presence/absence results are consistent with logistic regression

Cover:

-   More consistent results when looking at number of fish vs. presence/absence

```{r}

mark_gard_1998_hsi <- function(data, percent_presence = 20, cols = c()) {
  data_tidy <- data |> 
    select(count, cols) |> 
    mutate(fish_presence = ifelse(count > 0, 1, 0)) |> 
    pivot_longer(cols = cols, names_to = "type", values_to = "percent") |> 
    mutate(presence = ifelse(percent >= percent_presence, 1, 0)) |> 
    filter(presence == 1) |> 
    group_by(type) |>
    summarise(n_fish_presence = sum(as.numeric(fish_presence)),
              n_fish_total = sum(count)) |> 
    mutate(prop_fish_presence = n_fish_presence/sum(n_fish_presence),
           prop_fish_total = n_fish_total/sum(n_fish_total)) |> 
    ungroup()
  
     most_freq_total <- data_tidy |> 
      arrange(desc(prop_fish_total)) |> 
      slice(1) 
 
     most_freq_presence <- data_tidy |> 
      arrange(desc(prop_fish_presence)) |> 
      slice(1) 
  
  
  data_tidy_final <- data_tidy |> 
    mutate(hsc_presence = prop_fish_presence/most_freq_presence$prop_fish_presence,
           hsc_total = prop_fish_total/most_freq_total$prop_fish_total)
  
  return(data_tidy_final)

}

```

```{r}

mini_locations_raw <- read_csv(here::here("mini-snorkel-app", "mini_locations_raw.csv"))
mini_fish_raw <- read_csv(here::here("mini-snorkel-app","mini_fish_raw.csv"))

feather_cover <- mini_fish_raw |> 
  left_join(mini_locations_raw |> 
              distinct(location_table_id, date, channel_location, location)) |> 
  # create cover variables that are more consistent with strategic plan categories
  mutate(small_woody_debris = percent_small_woody_cover_inchannel,
        large_woody_debris = percent_large_woody_cover_inchannel,
         boulder = percent_boulder_substrate,
         cobble = percent_cobble_substrate, 
         undercut_bank = percent_undercut_bank, 
         aquatic_veg = percent_submerged_aquatic_veg_inchannel, 
         overhanging_veg = percent_cover_half_meter_overhead + percent_cover_more_than_half_meter_overhead) |> 
  select(micro_hab_data_tbl_id, location_table_id, transect_code, location,
         small_woody_debris:overhanging_veg, count, date) |>
  distinct() 
```

```{r}
cols <- c('small_woody_debris', 'large_woody_debris', 'boulder', 'cobble', 'undercut_bank', 'aquatic_veg', 'overhanging_veg')

mark_gard_1998_hsi(data = feather_cover, percent_presence = 20, cols = cols) |> 
  pivot_longer(cols = c(hsc_presence, hsc_total), names_to = "HSI", values_to = "values") |> 
  ggplot() + 
  geom_col(aes(y = values, x = type, fill = HSI), position = "dodge") +
  coord_flip() +
  ggtitle('Mark Gard 1998: HSI Comparison') + 
  theme_minimal()


```

```{r}

feather_cover |> 
    select(count, cols) |> 
    mutate(fish_presence = ifelse(count > 0, 1, 0)) |> 
    pivot_longer(cols = cols, names_to = "type", values_to = "percent") |> 
    mutate(presence = ifelse(percent >= 20, 1, 
                             ifelse(percent <20 & percent > 0, 0, NA))) |> 
    group_by(type, presence) |> 
    #filter(percent >= 20) |> 
    tally()
    
```

```{r}

feather_cover |> 
    select(count, cols) |> 
    mutate(fish_presence = ifelse(count > 0, 1, 0)) |> 
    pivot_longer(cols = cols, names_to = "type", values_to = "percent") |> 
    mutate(presence = ifelse(percent >= 20, 1, 
                             ifelse(percent <20 & percent > 0, 0, NA))) |> 
    group_by(type, presence) |> 
    filter(percent >= 20) |> 
    tally() |> 
    ungroup() |> 
    mutate(sum_n = sum(n),
           weighting_factor = 1-(n/sum_n))

```

### Questions/Next steps

-   How do HSI's compare across sites? Do we want an HSI per site?
-   Does it make sense to weight cover variables given some are just inherently less common (ie., undercut bank)

## Site Level 

Goal: compare
